### 2019-07-09 doc/nevh.tex
Initial commit

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[portuges]{babel}
\usepackage[T1]{fontenc}
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{amsmath,icomma,tikz,amsfonts}
\usepackage{bbold}

\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}

\title{\texttt{nevh}: Numerical Evolution from the Hamiltonian}
\date{\today}
\author{LJM~Amoreira}
\begin{document}
\maketitle
\section{Introduction}
This package computes the trajectories of mechanical systems given their
hamiltonian function and initial state alone.

The partial derivatives of the hamiltonian in order to the $q$'s and $p$'s are
estimated numerically and their values used to set up and solve the time ODE.

\section{Hamilton's Equations}
The physical state of a system with $N_f$ degreees of freedom is completely
specified given the values of $2N_f$ variables $q_i$ (coordinates) and $p_i$
(momentums) $i=1,\ldots,N_f$.  Its time evolution is determined by the
Hamilton's Equations,
\begin{align}
  \dot q_i&=\pd{H}{p_i}&
  \dot p_i&=-\pd{H}{q_i},
\end{align}
where $H=H(t,q,p)$ is the hamiltonian of the system which, in many situations,
is simply the toal mechanical energy of the system, $H=T+U$. Once the partial
derivatives in the rhs of Hamilton's equations are known, they define a system
of $2N_f$ first order simultaneous ordinary differential equations (ODE) in time
for the $q$'s and the $p$'s. The solution of this system is the trajectory (in
phase space) followed by the system in its evolution.

In most cases, the above mentioned system of ODEs must be solved numerically,
and a large number of computer packages, for all significant computer languages
have been developed to do that. Here, we take this numerical approach a step
further, and compute the partial derivatives of the hamiltonian also
numerically. With this approach, the evolution of the system can be obtained
from the hamiltonian alone, saving the need for the derivation of the right hand
sides of the Hamilton's equations.

\section{Hamilton's Equations in Numerical Form}
Hamilton's
Equations can be set in a more compact (and more suitable for numerical resolution
using standard routines or in the current approach) form,
\begin{equation}\label{eq:HEterse}
  \dot \psi=f(t,\psi),
\end{equation}
where $\psi=(\psi_1,\,\psi_2,\,\ldots,\,\psi_{2N_f})=(q_i,\,p_i)$,
$i=1,\ldots N_f$ and
\begin{equation}
  f(t,\psi)=
  \begin{pmatrix}
    \mathbb{0}&\mathbb{1}\\
    -\mathbb{1} &\mathbb{O}
  \end{pmatrix}\nabla_\psi H(t,\psi).
\end{equation}
Here, $\mathbb{0}$ and $\mathbb{1}$ represent the $N_f\times N_f$ zero matrix
and identity matrix, respectively, and
\begin{equation*}
  \nabla_\psi H=\left(
    \pd{H}{\psi_1},\,
    \pd{H}{\psi_2},\,\ldots,\,
    \pd{H}{\psi_{2N_f}}
    \right).
\end{equation*}
\end{document}

### 2019-07-09 src/hgrad.py

@@ -1,0 +1,18 @@
+def hgrad(H, t, psi, dpsi, **hparams):
+    ndf = int(len(gradient)/2)
+    i = np.eye(ndf)
+    z = np.zeros((ndf,ndf))
+    hswap = np.vstack(
+            (np.hstack((z,i)),
+             np.hstack((-i,z)))
+            )
+
+    gradient = np.array(
+            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
+              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
+             for i in range(n)])
+    hgradient = np.zeros(2*ndf)
+    hgradient[:ndf] = gradient[ndf:]
+    hgradient[ndf:] = -gradient[:ndf]
+    return hgradient
+

### 2019-07-09 src/nhev.py

@@ -1,0 +1,109 @@
+import numpy as np
+from copy import deepcopy
+
+
+def grad(H, t, psi, dpsi, **hparams):
+    """grad(H, t, psi, dpsi, **hparams)
+    Given function H(t, psi, **hparams), return numerical estimate of gradient
+    at t, psi, using step vector dpsi.
+    The gradient is computed using central differences. The returned gradient
+    has shape equal to psi.shape. 
+
+    Parameters:
+    ===========
+    H: callable with signature H(t, psi, *params)
+    t: double
+    psi: array of doubles storing qs and ps
+    dpsi: array of doubles (steps for finite difference formulas)
+    **hparams: other parameters of H
+
+    Returns:
+    ========
+    gradient: ndarray
+    
+    Examples:
+    =========
+    >>> grad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
+    array([1., 2.])
+    """
+    n = len(psi)
+    identity = np.eye(n)
+    return np.array(
+            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
+              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
+             for i in range(n)])
+
+
+def hgrad(H, t, psi, dpsi, **hparams):
+    """hgrad(H, t, psi, dpsi, **params)
+    Given function H(t, psi, **hparams), return numerical estimate of
+    "Hamilton's gradient", (dH/dpsi[1], -dH/dpsi[0])
+    The partial derivatives are computed using grad. hgrad has shape equal to
+    psi's.
+
+    Parameters:
+    ===========
+    H: callable with signature H(t, psi, *params)
+    t: double
+    psi: array of doubles storing qs and ps
+    dpsi: array of doubles (steps for finite difference formulas)
+    **hparams: other parameters of H
+
+    Returns:
+    ========
+    gradient: ndarray
+
+    Examples:
+    =========
+    >>> hgrad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
+    array([ 2., -1.])
+    """
+    ndf = int(len(gradient)/2)
+    i = np.eye(ndf)
+    z = np.zeros((ndf,ndf))
+    hswap = np.vstack(
+            (np.hstack((z,i)),
+             np.hstack((-i,z)))
+            )
+
+    gradient = np.array(
[…]

### 2019-08-12 doc/nevh.tex
Resuming work. Cleaned up a bit.

@@
 where $H=H(t,q,p)$ is the hamiltonian of the system which, in many situations,
-is simply the toal mechanical energy of the system, $H=T+U$. Once the partial
+is simply the total mechanical energy of the system, $H=T+U$. Once the partial
 derivatives in the rhs of Hamilton's equations are known, they define a system
@@
 \section{Hamilton's Equations in Numerical Form}
-Hamilton's
-Equations can be set in a more compact (and more suitable for numerical resolution
-using standard routines or in the current approach) form,
+Hamilton's Equations can be set in a more compact (and more suitable for
+numerical resolution using standard routines or in the current approach) form,
 \begin{equation}\label{eq:HEterse}
@@
 \end{equation*}
+This package
 \end{document}

### 2019-08-12 src/nevh.py

@@ -58,18 +58,18 @@ def hgrad(H, t, psi, dpsi, **hparams):
     >>> hgrad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
     array([ 2., -1.])
     """
-    ndf = int(len(gradient)/2)
-    i = np.eye(ndf)
-    z = np.zeros((ndf,ndf))
+    ndf = int(len(psi)/2)
+    I = np.eye(ndf)
+    Z = np.zeros((ndf,ndf))
     hswap = np.vstack(
-            (np.hstack((z,i)),
-             np.hstack((-i,z)))
+            (np.hstack((Z,I)),
+             np.hstack((-I,Z)))
             )
 
     gradient = np.array(
-            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
-              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
-             for i in range(n)])
+            [(H(t, psi + dpsi[i]*I[i]/2, **hparams)-
+              H(t, psi - dpsi[i]*I[i]/2, **hparams))/dpsi[i]
+             for i in range(ndf)])
     hgradient = np.zeros(2*ndf)
     hgradient[:ndf] = gradient[ndf:]
     hgradient[ndf:] = -gradient[:ndf]
@@ -77,7 +77,7 @@ def hgrad(H, t, psi, dpsi, **hparams):
 
 
 # Signature for hamiltonian functions: H(t, psi, **hparamss)
-# q, p -> arrays with dim = ndf (number of degrees of fereedom)
+# pi = (q, p) -> arrays with dim = ndf (number of degrees of fereedom)
 # t -> time
 # **hparams -> Other, problem-specific parameters in H
 #               (masses, elastic constants, g, etc)
@@ -106,4 +106,3 @@ def trajectory(H, psi0, ti, tf, nsteps, dpsi, **hparams):
         psin = psic + (k1 + 2*k2 + 2*k3 + k4) / 6
         traj[:,t_index+1] = np.copy(psin)
     return times, traj
-

### 2019-08-12 src/nevh.py
Working version.

@@ -11,7 +11,7 @@ def grad(H, t, psi, dpsi, **hparams):
 
     Parameters:
     ===========
-    H: callable with signature H(t, psi, *params)
+    H: callable with signature H(t, psi, **params)
     t: double
     psi: array of doubles storing qs and ps
     dpsi: array of doubles (steps for finite difference formulas)
@@ -34,67 +34,46 @@ def grad(H, t, psi, dpsi, **hparams):
              for i in range(n)])
 
 
-def hgrad(H, t, psi, dpsi, **hparams):
-    """hgrad(H, t, psi, dpsi, **params)
-    Given function H(t, psi, **hparams), return numerical estimate of
-    "Hamilton's gradient", (dH/dpsi[1], -dH/dpsi[0])
-    The partial derivatives are computed using grad. hgrad has shape equal to
-    psi's.
-
+def trajectory(H, psi0, ti, tf, nsteps, dpsi, **hparams):
+    """trajectory(H, s0, ti, tf, nsteps, ds, **hparams)
+    Solves Hamilton's equations for hamiltonian H and initial state s0 at nsteps
+    times in the range [ti, tf]. The partial derivatives are computed using
+    central derivation formulas with steps ds. 
+    H must be def'ed with signature H(t, s, **hparams)
+    
     Parameters:
     ===========
-    H: callable with signature H(t, psi, *params)
-    t: double
-    psi: array of doubles storing qs and ps
-    dpsi: array of doubles (steps for finite difference formulas)
+    H: callable with signature H(t, s, **params)
+    s0: array of doubles; initial state
+    ti, tf: double; initial and final times
+    nsteps: integer; number of times at wich the solutions is computed
+    ds: array of doubles (steps for finite difference formulas)
     **hparams: other parameters of H
 
     Returns:
     ========
-    gradient: ndarray
-
-    Examples:
-    =========
-    >>> hgrad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
-    array([ 2., -1.])
+    times: array of doubles; times at wich the solutions is computed
+    traj: array of doubles width shape (2*ndf, nsteps+1); state vetors at each
+          time in times
     """
-    ndf = int(len(psi)/2)
-    I = np.eye(ndf)
-    Z = np.zeros((ndf,ndf))
-    hswap = np.vstack(
[…]

### 2019-08-12 doc/nevh.tex
Cleaned up a bit, added some documentation.

@@
 \date{\today}
-\author{LJM~Amoreira}
+\author{JML~Amoreira\\
+  \texttt{jmlamoreira@gmail.com}
+  \and LJM~Amoreira\\
+  \texttt{amoreira@ubi.pt}
+}
 \begin{document}
@@
 \end{equation*}
-This package
+This package computes $\nabla_\psi H$ using central finite difference formulas
+and solves the motion equations using the plain RK4 method. The user only has to
+supply de definition of the hamiltonian and the initial state, no need to
+explicitly define the Hamilton's equations for the problem.
+
+Some simple examples are presented in the notebook in the \texttt{test/} folder.
 \end{document}

### 2019-08-12 src/nevh.py

@@ -1,5 +1,8 @@
+# 2019-08-12
+# JML Amoreira: jmlamoreira@gmail.com
+# LJM Amoreira: amoreira@ubi.pt, ljmamoreira@gmail.com
+
 import numpy as np
-from copy import deepcopy
 
 
 def grad(H, t, psi, dpsi, **hparams):

### 2019-08-12 src/hgrad.py
More cleaning.

@@ -1,18 +1,0 @@
-def hgrad(H, t, psi, dpsi, **hparams):
-    ndf = int(len(gradient)/2)
-    i = np.eye(ndf)
-    z = np.zeros((ndf,ndf))
-    hswap = np.vstack(
-            (np.hstack((z,i)),
-             np.hstack((-i,z)))
-            )
-
-    gradient = np.array(
-            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
-              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
-             for i in range(n)])
-    hgradient = np.zeros(2*ndf)
-    hgradient[:ndf] = gradient[ndf:]
-    hgradient[ndf:] = -gradient[:ndf]
-    return hgradient
-

### 2020-07-24 paper/paper.tex
Added paper.tex

\documentclass{iopart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\begin{document}
\title{Numerical Evolution from the Hamiltonian}
\author{José M L Amoreira$^1$ and Luís J M Amoreira$^2$}
\address{$^1$ Departamento de Física, Instituto Superior Técnico, Lisboa,
Portugal}
\address{$^2$ Departamento de Física, Universidade da Beira Interior, Covilhã,
Portugal}
\ead{amoreira@ubi.pt}
\begin{abstract}
  We propose a numerical method for the approximate calculation of the time
  evolution of point particle systems given only the system's hamiltonian and
  initial condition. The numerical method both generates \emph{and} solves
  the equations of motion numerically. 
\end{abstract}
\submitto{\EJP}
\maketitle
\section{Introduction}
Here's the thing: Why dont't these changes get added?

\end{document}

### 2020-07-24 paper/paper.tex
Some edits to paper.tex

@@
   initial condition. The numerical method both generates \emph{and} solves
-  the equations of motion numerically. 
+  the equations of motion numerically. For demonstration purposes, a working
+  im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and applied to
+  standard problems. The method may have some pedagogical merits but the
+  numerical effort of generating the equations of motion makes it unsuitable for
+  actual numerical solution of ``real'' problems with any but just a few degrees
+  of freedom.
 \end{abstract}
@@
 \section{Introduction}
-Here's the thing: Why dont't these changes get added?
 

### 2020-08-07 paper/paper.tex
More edits to paper.tex

@@
 \usepackage[english]{babel}
+\expandafter\let\csname equation*\endcsname\relax
+\expandafter\let\csname endequation*\endcsname\relax
+\usepackage{amsmath}
+\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
 \begin{document}
@@
   We propose a numerical method for the approximate calculation of the time
-  evolution of point particle systems given only the system's hamiltonian and
-  initial condition. The numerical method both generates \emph{and} solves
-  the equations of motion numerically. For demonstration purposes, a working
-  im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and applied to
-  standard problems. The method may have some pedagogical merits but the
-  numerical effort of generating the equations of motion makes it unsuitable for
-  actual numerical solution of ``real'' problems with any but just a few degrees
-  of freedom.
+  evolution of point particle systems given only the system's hamiltonian
+  function and initial condition. The numerical method both generates \emph{and}
+  solves the equations of motion numerically. For demonstration purposes, a
+  working im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and
+  applied to standard problems. The method may have some pedagogical merits but
+  the numerical effort of generating the equations of motion makes it unsuitable
+  for actual numerical solution of ``real'' problems with any but just a few
+  degrees of freedom.
 \end{abstract}
+\noindent{\it Keywords\/}: Classical particle dynamics,
+Hamilton's equations,
+Numerical methods
+
 \submitto{\EJP}
-\maketitle
+%\maketitle
 \section{Introduction}
+In the hamiltonian formulation of classical mechanics, the time evolutin of
+physical systems is governed by Hamilton's equations which, for point point
+particle systems, take the form
+\begin{align}
+  \dot q_i&=\pd{H}{p_i}&
+  \dot p_i&=-\pd{H}{q_i},
+\end{align}
+for $i=1,\;\ldots,\;N$ where $N$ is the number of degrees of freedom of the
+system (three times the number of particle for unconstrined, three dimensional
+systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
+respectively, and $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
+(for XXX systems, the hamiltonian function is simply the mechanical energy of
+the system). For discrete systems, these equations are a set of first order
+ordinary differential equations (ODE) in time, whose solutions for given initial
+conditions are the trajectories foolowed by the system's particles.
 
+In all but a small handfull of very well known problems, this system of ODEs
+must be solved numerically. The usual approach in such cases is to obtain
+explicit expressions for the partial derivatives of the hamiltonian function
+(the right hand sides of Hamilton's equations) and to code those expressions in
+functions which are then supplied to a standard ODE solver subroutine. In this
+work, we propose an alternative numerical method, where the user only supplies
+the hamiltonian function of the system, leaving its partial derivatives to be
+computed numerically by the ODE solver.
+
+This procedure is computationally more expensive, since many evaluations of the
+hamiltonian function are needed in order to estimate its partial derivatives at
+any moment in time, and so it really is not suitable to the analisys of complex
+systems with many coupled degreess of freedom. However, for simple systems it is
+very practical and, anyway, it is an interesting comcept in it own that we
+haven't yet seen exposed anywhere.
+
+\section{Numerical procedure}
+\section{Simple implementation in \texttt{python} and examples}
+\section{Conclusion}
+
+\section*{References}
+\begin{thebibliography}{99}
+  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
+  (1980)
+  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
+  Physics.} Springer (2012)
+\end{thebibliography}
 \end{document}

### 2020-09-14 paper/paper.tex
Separated grad from nevh

@@
 \usepackage[english]{babel}
+\usepackage{bbm}
 \expandafter\let\csname equation*\endcsname\relax
@@
 \newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
+\newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
 \begin{document}
@@
 In the hamiltonian formulation of classical mechanics, the time evolutin of
-physical systems is governed by Hamilton's equations which, for point point
-particle systems, take the form
+physical systems is governed by Hamilton's equations which, for point particle
+systems, take the form
 \begin{align}
@@
 \end{align}
-for $i=1,\;\ldots,\;N$ where $N$ is the number of degrees of freedom of the
-system (three times the number of particle for unconstrined, three dimensional
+for $i=1,\;\ldots,\;F$ where $F$ is the number of degrees of freedom of the
+system (three times the number of particles for unconstrined, three dimensional
 systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
@@
 ordinary differential equations (ODE) in time, whose solutions for given initial
-conditions are the trajectories foolowed by the system's particles.
+conditions are the trajectories folowed by the system's particles.
 
@@
 any moment in time, and so it really is not suitable to the analisys of complex
-systems with many coupled degreess of freedom. However, for simple systems it is
-very practical and, anyway, it is an interesting comcept in it own that we
-haven't yet seen exposed anywhere.
+systems with more than just a few coupled degreess of freedom. However, for
+simple systems it is very practical and, anyway, it is an interesting concept in
+it own that we haven't yet seen exposed anywhere.
 
 \section{Numerical procedure}
+%The standard, reasonable approach to the problem of finding the trajectories of
+%$N$ interacting point particles using Hamilton's equations is as follows: one
+%first derives analytical expressions for the partial derivatives $\partial
+%H/\partial q_i$, $\partial H/\partial p_j$ of the system's Hamilton function;
+%next, these expressions are programmed and the code is supplied to some
+%computational routine implementing a numerical algorithm for ODEs, like Euler,
+%Runge-Kutta, etc. Such routines are readily available for virtually all
+%programming languages and interactve computational environments.
+Standard numerical routines for solving systems of ODEs expect the problem to be
+formulated in the form
+\begin{equation}\label{eq:s_ode}
+  \frac{d\tilde{\psi}}{dt}=\tilde{G}(t,
+  \tilde{\psi})
+\end{equation}
+where $\tilde\psi$ is an array with the functions to be determined and $\tilde
+G$ is a function of time and $\tilde\psi$. In Hamilton's equations for a system
+of $F$ degrees of freedom, we have
+\begin{equation}
+  \tilde\psi=[q_1, \ldots, q_F, p_1, \ldots, p_F]
+\end{equation}
+and
+\begin{equation}
+  \tilde G(t, \tilde\psi)=
+  \begin{pmatrix}
+    \mathbbm{O}&\mathbbm{1}\\-\mathbbm{1}&\mathbbm{O}
+  \end{pmatrix}\tilde\nabla H
+\end{equation}
+where $\mathbbm{O}$ and $\mathbbm{1}$ are, respectively, the $F\times F$ null
+and identity matrices and $\tilde\nabla H$ is a $2F$ componet column array whose
+elements are the
+partial derivatives of the hamiltonian function in order to the coordinate and 
+conjugate momentum variables $q_i$ and $p_j$,
+\begin{equation}
+  \tilde\nabla H = 
+  \begin{pmatrix}
+    \partial H/\partial q_1\\
+    \vdots\\
+    \partial H/\partial p_F
+  \end{pmatrix}.
+\end{equation}
+
+Usually, expressions for the partial derivatives of the hamiltonian function are
+deduced analitically and programmed in suitable form for being repeatedly called
+from the computational ODE solver routine. That is, the rhs of eq.
+(\ref{eq:s_ode}) is explicitly supplied by the user. Instead, in the methhod now
+proposed, the user only supplies the hamiltonian function, and it's up to the
+numerical ODE solver to compute the hamiltonian's partial derivatives as needed.
+
 \section{Simple implementation in \texttt{python} and examples}

### 2020-09-14 src/grad.py

@@ -1,0 +1,39 @@
+# 2019-08-12
+# JML Amoreira: jmlamoreira@gmail.com
+# LJM Amoreira: amoreira@ubi.pt, ljmamoreira@gmail.com
+
+import numpy as np
+
+
+def grad(H, t, psi, dpsi, **hparams):
+    """grad(H, t, psi, dpsi, **hparams)
+    Given function H(t, psi, **hparams), return numerical estimate of gradient
+    at t, psi, using step vector dpsi.
+    The gradient is computed using central differences. The returned gradient
+    has shape equal to psi.shape. 
+
+    Parameters:
+    ===========
+    H: callable with signature H(t, psi, **params)
+    t: double
+    psi: array of doubles storing qs and ps
+    dpsi: array of doubles (steps for finite difference formulas)
+    **hparams: other parameters of H
+
+    Returns:
+    ========
+    gradient: ndarray
+    
+    Examples:
+    =========
+    >>> grad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
+    array([1., 2.])
+    """
+    n = len(psi)
+    identity = np.eye(n)
+    return np.array(
+            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
+              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
+             for i in range(n)])
+
+

### 2020-09-14 src/nevh.py

@@ -4,39 +4,6 @@
 
 import numpy as np
 
-
-def grad(H, t, psi, dpsi, **hparams):
-    """grad(H, t, psi, dpsi, **hparams)
-    Given function H(t, psi, **hparams), return numerical estimate of gradient
-    at t, psi, using step vector dpsi.
-    The gradient is computed using central differences. The returned gradient
-    has shape equal to psi.shape. 
-
-    Parameters:
-    ===========
-    H: callable with signature H(t, psi, **params)
-    t: double
-    psi: array of doubles storing qs and ps
-    dpsi: array of doubles (steps for finite difference formulas)
-    **hparams: other parameters of H
-
-    Returns:
-    ========
-    gradient: ndarray
-    
-    Examples:
-    =========
-    >>> grad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
-    array([1., 2.])
-    """
-    n = len(psi)
-    identity = np.eye(n)
-    return np.array(
-            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
-              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
-             for i in range(n)])
-
-
 def trajectory(H, psi0, ti, tf, nsteps, dpsi, **hparams):
     """trajectory(H, s0, ti, tf, nsteps, ds, **hparams)
     Solves Hamilton's equations for hamiltonian H and initial state s0 at nsteps
@@ -66,7 +33,9 @@ def trajectory(H, psi0, ti, tf, nsteps, dpsi, **hparams):
     times = np.linspace(ti, tf, nsteps + 1) # One more to store ti
     traj = np.zeros((2*ndf, nsteps + 1))    # One more to store S_i = (qi,pi)
     I = np.eye(ndim)
-    # Internal function, so that I is accessible
+    # Using an internal, inlined definition instead of grad, so that I is
+    # accessible, no need to calculate its dimensions and generate it each 
+    # time grad is called
     def hgrad(H, t, psi, dpsi, **hparams):
         gradient = np.array(
                 [(H(t, psi + dpsi[i]*I[i]/2, **hparams)-

### 2020-09-17 paper/paper.tex
Freeze before rewrite of paper's S2

@@
 \newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
+\newcommand{\hmatrix}{\ensuremath{%
+  \begin{pmatrix}
+    \mathbbm{O}&\mathbbm{I}\\-\mathbbm{I}&\mathbbm{O}
+  \end{pmatrix}
+}}
+
 \begin{document}
@@
 for $i=1,\;\ldots,\;F$ where $F$ is the number of degrees of freedom of the
-system (three times the number of particles for unconstrined, three dimensional
+system (three times the number of particles for unconstrained, three dimensional
 systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
@@
 
-In all but a small handfull of very well known problems, this system of ODEs
-must be solved numerically. The usual approach in such cases is to obtain
+In all but a small handfull of very well known simple problems, this system of
+ODEs must be solved numerically. The usual approach in such cases is to obtain
 explicit expressions for the partial derivatives of the hamiltonian function
 (the right hand sides of Hamilton's equations) and to code those expressions in
-functions which are then supplied to a standard ODE solver subroutine. In this
+subprograms which are then supplied to a standard ODE solver subroutine. In this
 work, we propose an alternative numerical method, where the user only supplies
 the hamiltonian function of the system, leaving its partial derivatives to be
-computed numerically by the ODE solver.
+computed numerically by the ODE solver itself.
 
@@
 hamiltonian function are needed in order to estimate its partial derivatives at
-any moment in time, and so it really is not suitable to the analisys of complex
+any moment in time, and so it really is not suitable to the analysis of complex
 systems with more than just a few coupled degreess of freedom. However, for
 simple systems it is very practical and, anyway, it is an interesting concept in
-it own that we haven't yet seen exposed anywhere.
+it own that we haven't seen exposed elsewhere.
 
+%-------------------------------------------------------------------------------
 \section{Numerical procedure}
@@
 \begin{equation}\label{eq:s_ode}
-  \frac{d\tilde{\psi}}{dt}=\tilde{G}(t,
-  \tilde{\psi})
+  \frac{d\tilde{\psi}}{dt}=\tilde{G}(t, \tilde{\psi})
 \end{equation}
@@
 of $F$ degrees of freedom, we have
-\begin{equation}
-  \tilde\psi=[q_1, \ldots, q_F, p_1, \ldots, p_F]
+\begin{equation}\label{eq:hpsi}
+  \tilde\psi(t)=[q_1(t), \ldots, q_F(t), p_1(t), \ldots, p_F(t)],\qquad
+  H=H(t,\tilde\psi)
 \end{equation}
 and
-\begin{equation}
+\begin{equation}\label{eq:heqsg}
   \tilde G(t, \tilde\psi)=
-  \begin{pmatrix}
-    \mathbbm{O}&\mathbbm{1}\\-\mathbbm{1}&\mathbbm{O}
-  \end{pmatrix}\tilde\nabla H
+  \hmatrix
+  \tilde\nabla H(t,\tilde{\psi})
 \end{equation}
-where $\mathbbm{O}$ and $\mathbbm{1}$ are, respectively, the $F\times F$ null
+where $\mathbbm{O}$ and $\mathbbm{I}$ are, respectively, the $F\times F$ null
 and identity matrices and $\tilde\nabla H$ is a $2F$ componet column array whose
@@
 from the computational ODE solver routine. That is, the rhs of eq.
-(\ref{eq:s_ode}) is explicitly supplied by the user. Instead, in the methhod now
-proposed, the user only supplies the hamiltonian function, and it's up to the
-numerical ODE solver to compute the hamiltonian's partial derivatives as needed.
+(\ref{eq:s_ode}) is explicitly supplied by the user. Instead, in the method now
+proposed, the user only supplies the code for the calculation of hamiltonian
+function, and it's up to the numerical ODE solver to compute the hamiltonian's
+partial derivatives as needed.
+
+Let us consider a concrete problem and ODE solver to better differenciate the
+two alternatives. Take the simple harmonic oscilator, solved using the explicit
+forward Euler method~\cite{sir:2010}. The hamiltonian function is
+\begin{equation}
+  H(q,p) = \frac{1}{2}kq^2+\frac{p^2}{2m},
+\end{equation}
+where $k$ is the elastic constant of the restoring force and $m$ is the mass of
+the oscilator.
+
+%The solution of the system~(\ref{eq:s_ode})
+%given~\eqref{eq:hpsi}
+%and~\eqref{eq:heqsg} is obtained using the explicit forward Euler
+%method~\cite{sir:2010} by iterative application of the formula
+%\begin{equation*}
+%  \tilde\psi(t+\delta t)=\tilde\psi(t)+\delta t
+%  \hmatrix\tilde\nabla H(t,\tilde\psi(t))
+%\end{equation*}
 
@@
   Physics.} Springer (2012)
+  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
+  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
+  Cambridge University Press (2007)
 \end{thebibliography}

### 2020-09-24 paper/paper.tex

@@
   \end{pmatrix}
+}}
+\newcommand{\numap}[1]{\ensuremath{%
+    \left\lfloor#1\right\rceil
 }}
@@
 \section{Introduction}
-In the hamiltonian formulation of classical mechanics, the time evolutin of
-physical systems is governed by Hamilton's equations which, for point particle
-systems, take the form
-\begin{align}
+In the hamiltonian formulation of classical mechanics~\cite{gol:1980}, the time
+evolution of physical systems is governed by Hamilton's equations which, for
+point particle systems, take the form
+\begin{align}\label{eq:heqs}
   \dot q_i&=\pd{H}{p_i}&
@@
 \end{align}
-for $i=1,\;\ldots,\;F$ where $F$ is the number of degrees of freedom of the
+for $i=1,\;\ldots,\;N,$ where $N$ is the number of degrees of freedom of the
 system (three times the number of particles for unconstrained, three dimensional
 systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
-respectively, and $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
+respectively, $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
 (for XXX systems, the hamiltonian function is simply the mechanical energy of
-the system). For discrete systems, these equations are a set of first order
+the system) and dotted symbols denote their total derivative with respect to
+time. For discrete systems, these equations are a set of first order
 ordinary differential equations (ODE) in time, whose solutions for given initial
@@
 \section{Numerical procedure}
-%The standard, reasonable approach to the problem of finding the trajectories of
-%$N$ interacting point particles using Hamilton's equations is as follows: one
-%first derives analytical expressions for the partial derivatives $\partial
-%H/\partial q_i$, $\partial H/\partial p_j$ of the system's Hamilton function;
-%next, these expressions are programmed and the code is supplied to some
-%computational routine implementing a numerical algorithm for ODEs, like Euler,
-%Runge-Kutta, etc. Such routines are readily available for virtually all
-%programming languages and interactve computational environments.
 Standard numerical routines for solving systems of ODEs expect the problem to be
-formulated in the form
-\begin{equation}\label{eq:s_ode}
-  \frac{d\tilde{\psi}}{dt}=\tilde{G}(t, \tilde{\psi})
+written in the form
+\begin{equation}\label{eq:numheqs}
+  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
 \end{equation}
-where $\tilde\psi$ is an array with the functions to be determined and $\tilde
-G$ is a function of time and $\tilde\psi$. In Hamilton's equations for a system
-of $F$ degrees of freedom, we have
-\begin{equation}\label{eq:hpsi}
-  \tilde\psi(t)=[q_1(t), \ldots, q_F(t), p_1(t), \ldots, p_F(t)],\qquad
-  H=H(t,\tilde\psi)
+where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
+equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
+G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
+the user as callable subprograms. The ODE
+solver routine automatically invokes these subprograms in order (as many times
+as required by the sspecific method and/or desired accuracy) to compute
+estimates of the unknown functions $\psi_i$.
+
+For Hamilton's equations, the $\psi_i$ are the coordinates and momenta, while
+the $G_i$ are the partial derivatives of the hamiltonian function, for which, in
+the standard approach, the user must derive expressions to be programmed.
+Instead, in the method now proposed, only the hamiltonian function is coded and
+the ODE solver is modified so that it repeatedly invokes that subprogram in
+order to compute the hamiltonian's partial derivatives numerically. Only then
+does it proceed to the estimation of the ODE solution.
+
+To make this discussion clearer, let's consider a specific problem, the
+classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
+the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
+a classical oscilator with mass $m$ and restoring force constant $k$ is 
+\begin{equation}
+  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
 \end{equation}
-and
-\begin{equation}\label{eq:heqsg}
-  \tilde G(t, \tilde\psi)=
-  \hmatrix
-  \tilde\nabla H(t,\tilde{\psi})
+Hamilton's equations~\eqref{eq:heqs} then read
+\begin{align}
+  \frac{dq}{dt}&=\frac{p}{m}&
+  \frac{dp}{dt}&=-kq,
+\end{align}
+wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
+$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
+by iterating the formulas
+\begin{align}
+  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
+  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
+\end{align}
+where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
+implements this method, and this the usual procedure. With this paper, we
+consider a method where explicit expressions for the hamiltonian's partial
+derivatives in the equations of motion are replaced by numerical estimates. In
+our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
+and~\eqref{eq:efe2} by
+\begin{align*}
+  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
+  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
+\end{align*}
+where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
+In order to make numerical aproximations to the partial derivatives of a
+function, one must evaluate it several times; at the very least, two evaluations
+are needed. For example, using the central difference method, the derivative is
+estimated as
+\begin{equation} 
+  \frac{\partial\psi}{\partial x} = 
+    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
+  o(\delta x)^2,
 \end{equation}
-where $\mathbbm{O}$ and $\mathbbm{I}$ are, respectively, the $F\times F$ null
-and identity matrices and $\tilde\nabla H$ is a $2F$ componet column array whose
-elements are the
-partial derivatives of the hamiltonian function in order to the coordinate and 
-conjugate momentum variables $q_i$ and $p_j$,
-\begin{equation}
-  \tilde\nabla H = 
-  \begin{pmatrix}
-    \partial H/\partial q_1\\
-    \vdots\\
-    \partial H/\partial p_F
-  \end{pmatrix}.
-\end{equation}
+leading to approximation formulas that, for our example problem, read 
+\begin{align}
+  \numap{\frac{\partial H}{\partial p}}_i &=
+  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
+  \numap{\frac{\partial H}{\partial q}}_i &=
+  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
+\end{align}
+where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
+Other approximations have higher accuracy, at the cost of extra evaluations of
+the function to be derived.
 
-Usually, expressions for the partial derivatives of the hamiltonian function are
-deduced analitically and programmed in suitable form for being repeatedly called
-from the computational ODE solver routine. That is, the rhs of eq.
-(\ref{eq:s_ode}) is explicitly supplied by the user. Instead, in the method now
-proposed, the user only supplies the code for the calculation of hamiltonian
-function, and it's up to the numerical ODE solver to compute the hamiltonian's
-partial derivatives as needed.
-
-Let us consider a concrete problem and ODE solver to better differenciate the
-two alternatives. Take the simple harmonic oscilator, solved using the explicit
-forward Euler method~\cite{sir:2010}. The hamiltonian function is
-\begin{equation}
-  H(q,p) = \frac{1}{2}kq^2+\frac{p^2}{2m},
-\end{equation}
-where $k$ is the elastic constant of the restoring force and $m$ is the mass of
-the oscilator.
-
-%The solution of the system~(\ref{eq:s_ode})
-%given~\eqref{eq:hpsi}
-%and~\eqref{eq:heqsg} is obtained using the explicit forward Euler
-%method~\cite{sir:2010} by iterative application of the formula
-%\begin{equation*}
-%  \tilde\psi(t+\delta t)=\tilde\psi(t)+\delta t
-%  \hmatrix\tilde\nabla H(t,\tilde\psi(t))
-%\end{equation*}
+This need for repeated evaluation of the system's hamiltonian reduces the
+efficiency of this method. However, programming a procedure for the hamiltonian
+function is usually much simpler then doing it for the set of its partial
+derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
+hamiltonian procedure is likewise much simpler than when the partial derivatives
+are coded.  For small systems, using the hamiltonian is much simpler and faster.
 
@@
   (1980)
+  \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
+  Company (1971)
   \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in

### 2020-09-24 src/nevh.py
Rewrote nevh to isolate from ODE solver. Rewrote examples notebook.

@@ -9,22 +9,24 @@
 import numpy as np
 
 class Hgrad():
-    def __init__(self, H, N, dpsi, **hparams):
+    def __init__(self, H, dpsi, **hparams):
         self.H = H
-        self.N = N
+        self.N = len(dpsi)
+        self.ndf = self.N // 2
         self.dpsi = dpsi
         self.hparams = hparams
-        self.I = np.eye(2 * N)
+        self.I = np.eye(self.N)
 
     def __call__(self, t, psi):
         delta_H = np.array(
             [self.H(t, psi + self.dpsi[i]*self.I[i]/2, **self.hparams)-
              self.H(t, psi - self.dpsi[i]*self.I[i]/2, **self.hparams)
-                 for i in range(2*self.N)])
+                 for i in range(self.N)])
         grad = delta_H / self.dpsi
-        hgrad = np.zeros(2*self.N)
-        hgrad[:self.N] = grad[self.N:]
-        hgrad[self.N:] = -grad[:self.N]
+        hgrad = np.zeros(self.N)
+        ndf = self.N // 2
+        hgrad[:self.ndf] = grad[self.ndf:]
+        hgrad[self.ndf:] = -grad[:self.ndf]
         return hgrad
 
 

### 2020-10-05 paper/paper.tex
Added to paper

@@
 G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
-the user as callable subprograms. The ODE
-solver routine automatically invokes these subprograms in order (as many times
-as required by the sspecific method and/or desired accuracy) to compute
-estimates of the unknown functions $\psi_i$.
+the user as callable subprograms. During its execution, the ODE solver routine
+repeatedly invokes these subprograms (as many times as required by the specific
+integration algorithm and/or desired accuracy) in order to compute estimates for
+the values the unknown functions $\psi_i$ at given values of $t$.
 
-For Hamilton's equations, the $\psi_i$ are the coordinates and momenta, while
-the $G_i$ are the partial derivatives of the hamiltonian function, for which, in
-the standard approach, the user must derive expressions to be programmed.
-Instead, in the method now proposed, only the hamiltonian function is coded and
-the ODE solver is modified so that it repeatedly invokes that subprogram in
-order to compute the hamiltonian's partial derivatives numerically. Only then
-does it proceed to the estimation of the ODE solution.
+For Hamilton's equations, the $\psi_i$ are the coordinates and conjugate momenta
+of the system, while the $G_i$ are the partial derivatives of its hamiltonian
+function. The usual procedure is to derive analitical expressions for the $G_i$
+as functions of $t$, the coordinates and the momenta, and then to code those
+expressions as subprograms to be called by the ODE solver routine.
 
-To make this discussion clearer, let's consider a specific problem, the
+In the method now proposed, instead, the user only has to supply code to
+compute the value of the system's hamiltonian function; its partial derivatives
+(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically by another
+subprogram called by the ODE solver.
+
+%is coded and
+%the ODE solver is modified so that it repeatedly invokes that subprogram in
+%order to compute the hamiltonian's partial derivatives numerically. Only then
+%does it proceed to the estimation of the ODE solution.
+
+To make this discussion clearer, let us consider a specific problem, the
 classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
@@
 function, one must evaluate it several times; at the very least, two evaluations
-are needed. For example, using the central difference method, the derivative is
-estimated as
-\begin{equation} 
+are needed. For example, using the second order central difference method, the
+derivative is estimated as
+\begin{equation}\label{eq:cdiff}
   \frac{\partial\psi}{\partial x} = 
@@
 
-\section{Simple implementation in \texttt{python} and examples}
+\section{Simple implementation in Python and examples}
+We made a simple implementation of this method in Python, using the standard
+numerical extensions library NumPy~\cite{numpy:2011,numpy:2020}.  It consists of
+a simple function that calls the hamiltonian function supplied by the user to
+evaluate its partial derivatives, using the central difference formulas of
+eqs.~\eqref{eq:cdiff}. This function is then supplied to NumPy's ODE solver
+\texttt{solve\_ivp}.  For greater efficiency, it is wrapped in a class, whose
+objects store initialization details on creation, like the hamiltonian's
+parameters, the number of degrees of freedom, the values of the discretization
+steps, etc. These objects are callable with signature adjusted to what the ODE
+solver expects.  The whole class definition fits in a few dozen lines of code
+and is very easy to use. It is available at
+
 \section{Conclusion}
@@
   (1980)
+%
   \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
   Company (1971)
+%
   \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
   Physics.} Springer (2012)
+%
   \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
@@
   Cambridge University Press (2007)
+%
+  \bibitem{numpy:2011} S.J.~van der Walt, S.C.~Colbert, G.~Varoquaux,
+  \textsl{The NumPy array: a structure for efficient numerical computation,}
+  Computing in Science \& Engineering, \textbf{13} no. 2, 22-30 (2011).
+  doi: 10.1109/MCSE.2011.37
+%
+  \bibitem{numpy:2020} C.R.~Harris, K.J.~Millman, S.J.~van der Walt et. al.,
+  \textsl{Array programming with NumPy,} Nature \textbf{585}, 357-362 (2020).
+  https://doi.org/10.1038/s41586-020-2649-2
 \end{thebibliography}

### 2021-03-01 paper/paper.tex
Freeze before restarting work on project.
Cleaned code; added docstrings.

@@
 \usepackage[english]{babel}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric}
 \usepackage{bbm}
@@
 \begin{abstract}
-  We propose a numerical method for the approximate calculation of the time
+  We propose a numerical method for approximate calculations of the time
   evolution of point particle systems given only the system's hamiltonian
-  function and initial condition. The numerical method both generates \emph{and}
-  solves the equations of motion numerically. For demonstration purposes, a
-  working im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and
-  applied to standard problems. The method may have some pedagogical merits but
-  the numerical effort of generating the equations of motion makes it unsuitable
-  for actual numerical solution of ``real'' problems with any but just a few
-  degrees of freedom.
+  function and initial conditions. The method both generates and solves the
+  equations of motion numerically. For demonstration purposes, a working
+  im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and applied
+  to standard problems. The method may have some pedagogical merits but the
+  numerical effort of generating the equations of motion makes it unsuitable for
+  actual numerical solution of ``real'' problems with any but just a few degrees
+  of freedom.
 \end{abstract}
@@
 \section{Introduction}
-In the hamiltonian formulation of classical mechanics~\cite{gol:1980}, the time
-evolution of physical systems is governed by Hamilton's equations which, for
-point particle systems, take the form
+%-----
+% 1 As equações de Hamilton de um sistema de massas pontuais formam um sistema
+% de N ODEs de 1ª ordem, com a forma 
+% \dot x_i= G_i(t, \tilde x)
+%-----
+In the hamiltonian formulation of classical mechanics~\cite{gol:1980, fw:2003},
+the time evolution of physical systems is governed by Hamilton's equations
+which, for point particle systems, take the form
 \begin{align}\label{eq:heqs}
   \dot q_i&=\pd{H}{p_i}&
-  \dot p_i&=-\pd{H}{q_i},
+  \dot p_i&=-\pd{H}{q_i},&
+  i&=1, 2, \ldots, N,
 \end{align}
-for $i=1,\;\ldots,\;N,$ where $N$ is the number of degrees of freedom of the
-system (three times the number of particles for unconstrained, three dimensional
-systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
-respectively, $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
-(for XXX systems, the hamiltonian function is simply the mechanical energy of
-the system) and dotted symbols denote their total derivative with respect to
-time. For discrete systems, these equations are a set of first order
-ordinary differential equations (ODE) in time, whose solutions for given initial
-conditions are the trajectories folowed by the system's particles.
+where $N$ is the number of degrees of freedom of the system (three times the
+number of particles for three dimensional, unconstrained systems), $q_i$, $p_i$
+are the canonical coordinates and conjugate momenta respectively, $H=H(t, q_i,
+p_i)$ is the hamiltonian function of the system (for systems with time
+independent potentials and constraints, the hamiltonian is simply their
+mechanical energy function~\cite{fw:2003}) and dotted symbols denote their
+total derivative with respect to time. 
+Renaming the canonical coordinates and momenta as
+\begin{equation}
+  x_j = 
+  \begin{cases}
+    q_i&\quad\text{if }j\leq N\\
+    p_{i-N}&\quad\text{if }j> N
+  \end{cases}
+  \qquad j=1, \ldots, 2N,
+\end{equation}
+Hamilton's equations take a more unified form
+\begin{equation}\label{eq:std_ode}
+  \dot x_j=G_j(t, x_1, x_2, \ldots, x_{2N}),\quad j=1, \ldots, 2N
+\end{equation}
+where
+\begin{equation}
+  G_j(t, x)=
+  \begin{cases}
+    \frac{\partial H}{\partial x_{j+N/2}},&\quad\text{if }j\leq N\\
+    \frac{\partial H}{\partial x_{j-N/2}},&\quad\text{if }j> N.
+  \end{cases}
+\end{equation}
+This expression can be further simplified as a matrix product:
+\begin{equation}\label{eq:rhsmf}
+  G_i=M_{i j}\frac{\partial H}{\partial x_j},\qquad
+  \text{with }
+  M=
+  \begin{pmatrix}
+    \mathbbm{O}&\mathbbm{I}\\
+    -\mathbbm{I}&\mathbbm{O}
+  \end{pmatrix},
+\end{equation}
+where $\mathbbm{O}$ and $\mathbbm{I}$ represent, respectively, the $N\times N$
+zero and identity matrices.  Equation~\eqref{eq:std_ode} shows the Hamilton's
+equations for a system of point particles with $N$ degrees of freedom as a
+system of $2N$ first order ordinary differential equations (ODEs) on $2N$
+variables $x_1, x_2, \ldots, x_{2N}$.
 
 In all but a small handfull of very well known simple problems, this system of
-ODEs must be solved numerically. The usual approach in such cases is to obtain
-explicit expressions for the partial derivatives of the hamiltonian function
-(the right hand sides of Hamilton's equations) and to code those expressions in
-subprograms which are then supplied to a standard ODE solver subroutine. In this
-work, we propose an alternative numerical method, where the user only supplies
-the hamiltonian function of the system, leaving its partial derivatives to be
-computed numerically by the ODE solver itself.
+ODEs must be solved numerically. All popular library routines for solving
+ODEs\footnote{Like ODEPACK \cite{odepack} for fortran, \texttt{odeint}
+  \cite{odeint} in c++ boost libraries or 
+  \texttt{scipy.integrate.solve\_ivp}~\cite{scipy} for python.}
+require the user to supply subprograms to compute the rhs functions $G_i$.
+The ODE solver invokes these subprograms to compute estimates for the
+values of the unknowns $x_i$ for arbitrary time $t$, given their values at a
+given instant $t_0$.
 
-This procedure is computationally more expensive, since many evaluations of the
-hamiltonian function are needed in order to estimate its partial derivatives at
-any moment in time, and so it really is not suitable to the analysis of complex
-systems with more than just a few coupled degreess of freedom. However, for
-simple systems it is very practical and, anyway, it is an interesting concept in
-it own that we haven't seen exposed elsewhere.
+In this work, we propose an alternative numerical method, where the user only
+supplies the hamiltonian function of the system, leaving its partial derivatives
+to be computed numerically as well.  This procedure is computationally more
+expensive, since many evaluations of the hamiltonian function are needed in
+order to estimate its partial derivatives at any moment in time, and so it
+really is not suitable for the analysis of complex systems with more than just a
+few coupled degreess of freedom. However, for simple systems it is very
+practical and, anyway, it is an interesting concept in its own, which we haven't
+seen exposed elsewhere.
 
@@
 G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
-the user as callable subprograms. During its execution, the ODE solver routine
-repeatedly invokes these subprograms (as many times as required by the specific
-integration algorithm and/or desired accuracy) in order to compute estimates for
-the values the unknown functions $\psi_i$ at given values of $t$.
+the user as callable subprograms. For instance, the python standard initial
+value problem solver routine has interface
+\texttt{solve\_ivp(rhs, time\_interval, psi\_0)}; here, \texttt{rhs} is the name
+of a phython function to compute the values of the rhs functions $G_i$.
+During its execution, the ODE solver routine repeatedly invokes these
+subprograms (as many times as required by the specific integration algorithm
+and/or desired accuracy) in order to compute estimates for the values the
+unknown functions $\psi_i$ at given values of $t$.
 
@@
 of the system, while the $G_i$ are the partial derivatives of its hamiltonian
-function. The usual procedure is to derive analitical expressions for the $G_i$
-as functions of $t$, the coordinates and the momenta, and then to code those
-expressions as subprograms to be called by the ODE solver routine.
+function (reordered and eventually symmetrized, according to
+eq.~\ref{eq:rhsmf}). The usual procedure is to derive analytical expressions
+for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
+code those expressions as subprograms to be called by the ODE solver routine.
 
@@
 compute the value of the system's hamiltonian function; its partial derivatives
-(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically by another
-subprogram called by the ODE solver.
+(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically and reordered by
+another subprogram which the ODE solver calls. Note that the differentiation and
+reordering of the partial derivatives can be computed by a general purpose
+routine. Indeed, the gradient can even be computed also by a standard numerical
+library function, if available for the chosen computing system or language.
+Figure~\ref{fig:a} displays flowcharts for both methods.
+\begin{figure}[htb]
+  \centering
+    \begin{tikzpicture}[node distance=1.0cm, font=\sffamily,scale=0.7]
+      \footnotesize
+      \tikzstyle{io} = [trapezium,
+                        trapezium left angle=70,
+                        trapezium right angle=110,
+                        minimum width=1.0cm,
+                        minimum height=.6cm,
+                        text centered,
+                        draw=black,
+                        fill=blue!20]
+      \tikzstyle{libcode} = [rectangle,
+                             rounded corners,
+                             minimum width=1.5cm,
+                             text width=1.5cm,
+                             minimum height=0.6cm,
+                             text centered,
+                             draw=black,
+                             fill=orange!20]
+      \tikzstyle{usercode} = [rectangle,
+                             %rounded corners,
+                             minimum width=1.5cm,
+                             minimum height=0.6cm,
+                             text centered,
+                             draw=black,
+                             text width=1.5cm,
+                             fill=green!20]
+      \tikzstyle{arrow} = [thick,->,>=stealth]
 
-%is coded and
-%the ODE solver is modified so that it repeatedly invokes that subprogram in
-%order to compute the hamiltonian's partial derivatives numerically. Only then
-%does it proceed to the estimation of the ODE solution.
+      \begin{scope}[xshift=-3.5cm]
+        \node at (0,1) {Standard method};
+        \node (input) [io] {Initial state};
+        \node (odesolver) [libcode, below of=input] {ODE Solver};
+        \node (output) [io, below of=odesolver] {Final state};
+        \node (rhs) [usercode, right of=odesolver, node distance=2.3cm]
+          {$M\,\text{grad}\,H$};
+        \draw [arrow] (input) -- (odesolver);
+        \draw [arrow] (odesolver) -- (output);
+        \draw [arrow] (odesolver) -- (rhs);
+        \draw [arrow]  (rhs) -- (odesolver);
+      \end{scope}
 
-To make this discussion clearer, let us consider a specific problem, the
-classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
-the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
-a classical oscilator with mass $m$ and restoring force constant $k$ is 
-\begin{equation}
-  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
-\end{equation}
-Hamilton's equations~\eqref{eq:heqs} then read
-\begin{align}
-  \frac{dq}{dt}&=\frac{p}{m}&
-  \frac{dp}{dt}&=-kq,
-\end{align}
-wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
-$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
-by iterating the formulas
-\begin{align}
-  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
-  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
-\end{align}
-where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
-implements this method, and this the usual procedure. With this paper, we
-consider a method where explicit expressions for the hamiltonian's partial
-derivatives in the equations of motion are replaced by numerical estimates. In
-our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
-and~\eqref{eq:efe2} by
-\begin{align*}
-  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
-  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
-\end{align*}
-where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
-In order to make numerical aproximations to the partial derivatives of a
-function, one must evaluate it several times; at the very least, two evaluations
-are needed. For example, using the second order central difference method, the
-derivative is estimated as
-\begin{equation}\label{eq:cdiff}
-  \frac{\partial\psi}{\partial x} = 
-    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
-  o(\delta x)^2,
-\end{equation}
-leading to approximation formulas that, for our example problem, read 
-\begin{align}
-  \numap{\frac{\partial H}{\partial p}}_i &=
-  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
-  \numap{\frac{\partial H}{\partial q}}_i &=
-  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
-\end{align}
-where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
-Other approximations have higher accuracy, at the cost of extra evaluations of
-the function to be derived.
+      \begin{scope}[xshift=3.5cm]
+        \node at (0,1) {Proposed method};
+        \node (input) [io] {Initial state};
+        \node (odesolver) [libcode, below of=input] {ODE Solver};
+        \node (output) [io, below of=odesolver] {Final state};
+        \draw [arrow] (input) -- (odesolver);
+        \draw [arrow] (odesolver) -- (output);
+        \node (rhs) [libcode, right of=odesolver, node distance=2.5cm]
+          {General purpose\\ $M\,\text{grad}\,f$ routine};
+        \draw [arrow] (odesolver) -- (rhs);
+        \draw [arrow]  (rhs) -- (odesolver);
+        \node (h) [usercode, right of=rhs, node distance=2.5cm] {$H$ function};
+        \draw [arrow] (rhs) -- (h);
+        \draw [arrow] (h) -- (rhs);
+      \end{scope}
+    \end{tikzpicture}
+  \caption{Flowcharts for the standard and the proposed methods. Round cornered
+    boxes represent library or general purpose code; right angled boxes contain
+    code for the specific calculation at hand, which the user must
+  supply.\label{fig:a}}
+\end{figure}
 
-This need for repeated evaluation of the system's hamiltonian reduces the
-efficiency of this method. However, programming a procedure for the hamiltonian
-function is usually much simpler then doing it for the set of its partial
-derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
-hamiltonian procedure is likewise much simpler than when the partial derivatives
-are coded.  For small systems, using the hamiltonian is much simpler and faster.
+%To make this discussion clearer, let us consider a specific problem, the
+%classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
+%the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
+%a classical oscilator with mass $m$ and restoring force constant $k$ is 
+%\begin{equation}
+%  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
+%\end{equation}
+%Hamilton's equations~\eqref{eq:heqs} then read
+%\begin{align}
+%  \frac{dq}{dt}&=\frac{p}{m}&
+%  \frac{dp}{dt}&=-kq,
+%\end{align}
+%wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
+%$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
+%by iterating the formulas
+%\begin{align}
+%  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
+%  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
+%\end{align}
+%where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
+%implements this method, and this the usual procedure. With this paper, we
+%consider a method where explicit expressions for the hamiltonian's partial
+%derivatives in the equations of motion are replaced by numerical estimates. In
+%our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
+%and~\eqref{eq:efe2} by
+%\begin{align*}
+%  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
+%  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
+%\end{align*}
+%where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
+%In order to make numerical aproximations to the partial derivatives of a
+%function, one must evaluate it several times; at the very least, two evaluations
+%are needed. For example, using the second order central difference method, the
+%derivative is estimated as
+%\begin{equation}\label{eq:cdiff}
+%  \frac{\partial\psi}{\partial x} = 
+%    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
+%  o(\delta x)^2,
+%\end{equation}
+%leading to approximation formulas that, for our example problem, read 
+%\begin{align}
+%  \numap{\frac{\partial H}{\partial p}}_i &=
+%  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
+%  \numap{\frac{\partial H}{\partial q}}_i &=
+%  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
+%\end{align}
+%where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
+%Other approximations have higher accuracy, at the cost of extra evaluations of
+%the function to be derived.
 
+%This need for repeated evaluation of the system's hamiltonian reduces the
+%efficiency of this method. However, programming a procedure for the hamiltonian
+%function is usually much simpler then doing it for the set of its partial
+%derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
+%hamiltonian procedure is likewise much simpler than when the partial derivatives
+%are coded.  For small systems, using the hamiltonian is much simpler and faster.
+%
+%
+%
 \section{Simple implementation in Python and examples}
@@
 solver expects.  The whole class definition fits in a few dozen lines of code
-and is very easy to use. It is available at
+and is very easy to use. It is available for download as free software at
+github~\cite{nevh:2020}. The repository also includes a \texttt{jupyter
+notebook} showcasing several example applications.
 
@@
   (1980)
+%
+  \bibitem{fw:2003} 
+    A.L.~Fetter, J.D.~Walecka,
+    \textsl{Theoretical Mechanics of Particles and Continua.}
+    Dover (2003)
+%
+  \bibitem{odepack}
+  A.C. Hindmarsh, \textsl{ODEPACK, A Systematized Collection of ODE Solvers,} in
+    Scientific Com\-puting, R. S. Stepleman et al. (eds.), North-Holland,
+    Amsterdam, 1983 (vol. 1 of IMACS Transactions on Scientific Computation),
+    pp. 55-64. 
+%
+  \bibitem{odeint}
+  K. Ahnert and M. Mulansky, \textsl{Odeint - Solving Ordinary Differential
+  Equations in C++,} AIP Conf. Proc. 1389, pp. 1586-1589 (2011);
+doi:http://dx.doi.org/10.1063/1.3637934
+%
+  \bibitem{scipy}
+  P. Virtanen et al, \textsl{SciPy 1.0: Fundamental Algorithms for Scientific
+  Computing in Python}, Nature Methods 17, pp. 261--272 (2020)
 %
@@
   https://doi.org/10.1038/s41586-020-2649-2
+%
+  \bibitem{nevh:2020} J.M.L.~Amoreira, L.J.M.~Amoreira,
+  \texttt{https://github.com/ljmamoreira/nevh} (2020)
+
 \end{thebibliography}

### 2021-03-01 src/nevh.py

@@ -8,25 +8,64 @@
 
 import numpy as np
 
+# Hgrad: Hamiltonian grandient. 
 class Hgrad():
+    """
+    A class to compute the hamiltonian gradient of hamiltonian funcs.
+
+    Atributes
+    ---------
+    H: function H(t, [q_1,...,q_N, p_1,...,p_N])
+        Hamiltonian of physical system
+    N: int
+        number of degrees of freedom
+    dpsi: array[2N] of floats
+        step sizes for partial derivatives of hamiltonian
+    hparams: array[] of floats
+        parameters of the hamiltonian function
+    I: array[2N, 2N]
+        2Nx2N identity matrix
+
+    Methods
+    -------
+    __call__(t, np.array([q_1,...q_N, p_1,...p_N])
+        Returns a 2N np.array with the components of the hamiltonian gradient
+        [dH/dp_1,...,dH/dp_N, -dH/dq_1,...,-dH/dq_N]
+
+    """
+
     def __init__(self, H, dpsi, **hparams):
+        """
+        Creates and inits Hgrad objects
+
+        Parameters
+        ----------
+        H: function H(t, [q_1,...,q_N, p_1,...,p_N])
+            Hamiltonian of physical system
+        dpsi: array[2N] of floats
+            step sizes for partial derivatives of hamiltonian
+        hparams: array[] of floats
+            parameters of the hamiltonian function
+
+        """
+        # System global parameters  
         self.H = H
-        self.N = len(dpsi)
-        self.ndf = self.N // 2
-        self.dpsi = dpsi
-        self.hparams = hparams
-        self.I = np.eye(self.N)
+        self.N = len(dpsi) // 2     # Number of degrees of freedom
+        self.dpsi = dpsi            # Step sizes for differentiation
+        self.hparams = hparams      # Hamiltonian parameters
+        self.I = np.eye(2*self.N)   # Identity matrix
 
     def __call__(self, t, psi):
+        # Phase space gradient 
         delta_H = np.array(
             [self.H(t, psi + self.dpsi[i]*self.I[i]/2, **self.hparams)-
              self.H(t, psi - self.dpsi[i]*self.I[i]/2, **self.hparams)
-                 for i in range(self.N)])
+                 for i in range(2*self.N)])
         grad = delta_H / self.dpsi
-        hgrad = np.zeros(self.N)
-        ndf = self.N // 2
[…]

### 2021-10-13 paper/paper.tex
Freeze before introducing fractional steps for derivatives.

@@
 \usepackage{bbm}
+\usepackage{listings, algorithm2e}
+%\usepackage{xcolor}
+\lstset{%
+  language=python,
+  %backgroundcolor=\color{black!5},
+  basicstyle=\sffamily\footnotesize,
+  keepspaces=false,
+  showspaces=false,
+  showstringspaces=false
+}
 \expandafter\let\csname equation*\endcsname\relax
@@
 p_i)$ is the hamiltonian function of the system (for systems with time
-independent potentials and constraints, the hamiltonian is simply their
+independent potentials and constraints, the hamiltonian is simply the
 mechanical energy function~\cite{fw:2003}) and dotted symbols denote their
@@
   \begin{cases}
-    q_i&\quad\text{if }j\leq N\\
-    p_{i-N}&\quad\text{if }j> N
+    q_j&\quad\text{if }j\leq N\\
+    p_{j-N}&\quad\text{if }j> N
   \end{cases}
@@
   G_j(t, x)=
-  \begin{cases}
-    \frac{\partial H}{\partial x_{j+N/2}},&\quad\text{if }j\leq N\\
-    \frac{\partial H}{\partial x_{j-N/2}},&\quad\text{if }j> N.
+  \begin{cases}\displaystyle
+    \pd{H}{x_{j+N/2}} ,&\quad\text{if }j\leq N\\[1em]
+    \displaystyle
+    -\pd{H}{x_{j-N/2}}, &\quad\text{if }j> N.
   \end{cases}
@@
 \begin{equation}\label{eq:rhsmf}
-  G_i=M_{i j}\frac{\partial H}{\partial x_j},\qquad
+  G_i=M_{i j}\pd{H}{x_j},\qquad
   \text{with }
@@
 ODEs must be solved numerically. All popular library routines for solving
-ODEs\footnote{Like ODEPACK \cite{odepack} for fortran, \texttt{odeint}
+ODEs\footnote{Like ODEPACK~\cite{odepack} for fortran, \texttt{odeint}
   \cite{odeint} in c++ boost libraries or 
-  \texttt{scipy.integrate.solve\_ivp}~\cite{scipy} for python.}
+  \texttt{solve\_ivp} in SciPy~\cite{scipy} for python.}
 require the user to supply subprograms to compute the rhs functions $G_i$.
@@
 values of the unknowns $x_i$ for arbitrary time $t$, given their values at a
-given instant $t_0$.
+particular instant $t_0$.
 
@@
 equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
-G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
+G_i(t,\psi_j)$ are the rhs of the differential equations, which must be coded by
 the user as callable subprograms. For instance, the python standard initial
@@
 of the system, while the $G_i$ are the partial derivatives of its hamiltonian
-function (reordered and eventually symmetrized, according to
-eq.~\ref{eq:rhsmf}). The usual procedure is to derive analytical expressions
+function (reordered and eventually multiplied by -1, according
+to~(\ref{eq:rhsmf})). The usual procedure is to derive analytical expressions
 for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
@@
 compute the value of the system's hamiltonian function; its partial derivatives
-(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically and reordered by
+(the $G_i$ in~\eqref{eq:numheqs}) are estimated numerically and reordered by
 another subprogram which the ODE solver calls. Note that the differentiation and
@@
         \draw [arrow] (odesolver) -- (output);
-        \node (rhs) [libcode, right of=odesolver, node distance=2.5cm]
+        \node (rhs) [libcode, fill=yellow!35, right of=odesolver,
+                     node distance=2.5cm]
           {General purpose\\ $M\,\text{grad}\,f$ routine};
@@
 
+The procedure for the calculation of the partial derivatives of the Hamiltonian
+(the "General purpose $M\text{grad} f$" block in the flowchart to the right in
+Figure~\ref{fig:a}) can be implemented in many standard ways. For example, using
+central difference formulas, we can write it in pseudocode as\\
+\begin{algorithm}[H]
+  \For{$1\leq i\leq N$}{
+    $\partial_{q_i}H=\left(H(t,\; \tilde q, q_i+\delta q_i,\; \tilde{p})-
+    H(t,\; \tilde q, q_i-\delta q_i,\; \tilde{p})\right)/(2\delta q_i)$\\
+    $\partial_{p_i}H=\left(H(t,\; \tilde q,\; \tilde{p}, p_i+\delta p_i)-
+    H(t,\; \tilde q,\; \tilde{p}, p_i-\delta p_i)\right)/(2\delta p_i) $
+  } 
+  \Return
+  \{
+    $\partial_{p_1}H,\; \partial_{p_2}H,\;\ldots, \partial_{p_N}H,\;
+    -\partial_{q_1}H,\; -\partial_{q_2}H,\;\ldots, -\partial_{q_N}H$
+  \}
+
+\end{algorithm}
+\noindent
+where the notation $\tilde q$ above represents all the coordinates
+$q_1, \ldots, q_N$, except the $i$-th coordinate $q_i$, and similarly for the
+momenta ($\tilde p$).
+
 %To make this discussion clearer, let us consider a specific problem, the
-%classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
+%classical harmonic oscillator~\cite{french:71}, and a specific ODE solver method,
 %the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
-%a classical oscilator with mass $m$ and restoring force constant $k$ is 
+%a classical oscillator with mass $m$ and restoring force constant $k$ is 
 %\begin{equation}
@@
 %\end{align}
-%wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
+%which are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
 %$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
@@
 %
-\section{Simple implementation in Python and examples}
-We made a simple implementation of this method in Python, using the standard
-numerical extensions library NumPy~\cite{numpy:2011,numpy:2020}.  It consists of
-a simple function that calls the hamiltonian function supplied by the user to
-evaluate its partial derivatives, using the central difference formulas of
-eqs.~\eqref{eq:cdiff}. This function is then supplied to NumPy's ODE solver
-\texttt{solve\_ivp}.  For greater efficiency, it is wrapped in a class, whose
-objects store initialization details on creation, like the hamiltonian's
-parameters, the number of degrees of freedom, the values of the discretization
-steps, etc. These objects are callable with signature adjusted to what the ODE
-solver expects.  The whole class definition fits in a few dozen lines of code
-and is very easy to use. It is available for download as free software at
-github~\cite{nevh:2020}. The repository also includes a \texttt{jupyter
-notebook} showcasing several example applications.
+\section{Simple implementation in Python and an example}
+We made a simple implementation of this method in Python, using standard
+libraries like the numerical extensions package
+NumPy~\cite{numpy:2011,numpy:2020} and the scientific library
+SciPy~\cite{scipy}.  It consists of a simple function that calls the hamiltonian
+function supplied by the user to evaluate, reorder and return its partial
+derivatives, using central difference formulas. The step parameters $\delta q_i,
+\delta p_i$ for the derivation formulas must be supplied by the user.  This
+function is then given as an argument to SciPy's ODE solver \texttt{solve\_ivp}.
+For greater simplicity of use, it is wrapped in a class, whose objects store
+initialization details on creation, like the hamilton function, its parameters,
+the number of degrees of freedom, the values of the discretization steps, etc.
+Furthermore, these objects are defined as callable, returning the list of rhs of
+hamilton's equations for given values of the dynamical variables $q_i, p_i$ and
+time, so that they can be supplied to general ODE integrators just as if those
+rhs functions had been explicitly programmed.
+
+The whole class definition (including comments and python doc strings) fits in a
+few dozen lines of code and is very easy to use. It is available for download as
+free software at github~\cite{nevh:2020}.  The repository also includes a
+\texttt{jupyter notebook} showcasing several example applications.  Here we
+present a particularly simple application example, the one-dimensional harmonic
+oscillator (also included in the repository).
+
+\begin{lstlisting}[caption= {Numerical solution of Hamiltonian's equations for
+the one-dimensional linear harmonic oscillator. Note that the user only supplies
+code for the system's Hamilton function in lines 7-9. The remaining code lines
+define initial state, physical and numerical parameters.},
+numbers=left, stepnumber=1, label={lst:1dlho}, language=python
+]
+import numpy as np
+from scipy.integrate import solve_ivp
+import matplotlib.pyplot as plt
+import nevh
+
+# Hamiltonian function
+def H(t, s, k, m):
+    x, p = s
+    return 0.5*k*x**2 + 0.5*p**2/m
+
+# "Administrative" details
+# Initial state: off equilibrium position, at rest
+x0 = 1.0; p0 = 0.0; initial_state = [x0,p0]
+# Integration steps parameters for partial derivatives dH/dq and dH/dp
+ds = np.ones(2)*0.1
+# Hamiltonian parameters. With k=4\pi^2, m=1, the period is 1
+kc = 4 * np.pi**2; mc = 1.0
+wc = (kc / mc)**0.5; Tc = 2 * np.pi / wc
+
+# Define object that computes and returns the rhs of Hamilton's eqs
+G = nevh.Hgrad(H, ds, k=kc, m=mc)
+tmin = 0; tmax = 2 * Tc # two full oscillation periods
+# Solve Hamilton's equations
+trj = solve_ivp(G, [tmin, tmax], initial_state)
+\end{lstlisting}
+Scipy's ODE solver function \texttt{solve\_ivp} returns a composite structure
+(named \texttt{trj} in Listing~\ref{lst:1dlho}) that stores information
+regarding the numerical solution of the ODE. Namely, sampled values of the
+coordinates and momenta at different times in the interval \texttt{[tmin, tmax]}
+are stored at the arrays \texttt{trj.y[0]}, \texttt{trj.y[1]}, while the
+sampling times are stored in \texttt{trj.t}. The sampling frequency can be
+adjusted by the user, but here we just took the default behaviour of
+\texttt{solve\_ivp}, which sets that frequency accordind to precision
+requirements (that we also didn't bother to specify, because it wasn't necessary
+for convergence).
+
+More relevant to the method here presented, the values of the discretization
+steps used for the numerical computation of the partial derivatives of the
+hamiltonian, stored in array \texttt{ds} in Listing~\ref{lst:1dlho}, must be
+supplied by the user. In the example shown these values are irrelevant, because
+the hamiltonian of the linear harmonic oscilator is a quadratic function of all
+the coordinates and conjugate momenta and the central difference formulas
+yield exact estimates of the derivatives of quadratic functions. However, in
+other problems, the values of the discretizations steps must be carefully
+considered.
+
+Figure~\ref{fig:1dlho}
+displays the graphs of coordinates and momenta of the numerical solution
+obtained (discrete points), together with plots of the analytical solutions for
+the same problem.
+\begin{figure}[htb]
+  \begin{center}
+    \includegraphics[width=0.6\linewidth]{figs/1dlho.png}
+  \end{center}
+  \caption{Plot of position (\texttt{x}) and momentum (\texttt{p}) for the
+  one-dimensional harmonic oscillator problem of Listing~\ref{lst:1dlho}.
+  }\label{fig:1dlho}
+\end{figure}
+The correctness of the numerical approximation, at least for pedagogical
+purposes only, is manifest.
+
+Note that we do not claim that this method is more efficient than the standard
+approach. Much to the contrary, in computationally intensive applications, the
+effort involved in the numerical calculation of the partial derivatives of the
+hamiltonian (for which several computations of the hamiltonian function itself
+are needed), at every instant in the simulation, slows the calculaton very
+significantly. What we do claim is that our method makes it very simple to set
+up a practical simulation of small mechanical systems, because the user is
+spared the trouble of deriving expressions for the partial derivatives of the
+hamiltonian and writing subroutines that compute them, with the calling
+signature that the particular ODE solver expects.
 

### 2021-10-13 src/nevh.py

@@ -68,7 +68,3 @@ class Hgrad():
         hgrad[self.N:] = -grad[:self.N]
         return hgrad
 
-
-
-
-

### 2022-09-07 paper/paper.tex
Freeze before separating paper

--- 2021-03-01 paper/paper.tex
@@
 \usepackage{bbm}
+\usepackage{listings, algpseudocode, amssymb}
+%\usepackage{xcolor}
+\lstset{%
+  language=python,
+  %backgroundcolor=\color{black!5},
+  basicstyle=\sffamily\footnotesize,
+  keepspaces=false,
+  showspaces=false,
+  showstringspaces=false
+}
 \expandafter\let\csname equation*\endcsname\relax
@@
   We propose a numerical method for approximate calculations of the time
-  evolution of point particle systems given only the system's hamiltonian
+  evolution of point particle systems given only the system's Hamiltonian
   function and initial conditions. The method both generates and solves the
   equations of motion numerically. For demonstration purposes, a working
-  im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and applied
+  im\-ple\-men\-ta\-tion written in Python is described and applied
   to standard problems. The method may have some pedagogical merits but the
@@
 %-----
-In the hamiltonian formulation of classical mechanics~\cite{gol:1980, fw:2003},
+In the Hamiltonian formulation of classical mechanics~\cite{gol:1980, fw:2003},
 the time evolution of physical systems is governed by Hamilton's equations
@@
 are the canonical coordinates and conjugate momenta respectively, $H=H(t, q_i,
-p_i)$ is the hamiltonian function of the system (for systems with time
-independent potentials and constraints, the hamiltonian is simply their
+p_i)$ is the Hamiltonian function of the system (for systems with time
+independent potentials and constraints, the Hamiltonian is simply the
 mechanical energy function~\cite{fw:2003}) and dotted symbols denote their
-total derivative with respect to time. 
-Renaming the canonical coordinates and momenta as
-\begin{equation}
-  x_j = 
+total derivatives with respect to time. 
+Renaming the canonical coordinates $q_i$ and momenta $p_i$ as
+\begin{equation}\label{eq:xinot}
+  \nu_i = 
   \begin{cases}
-    q_i&\quad\text{if }j\leq N\\
-    p_{i-N}&\quad\text{if }j> N
+    q_i&\quad\text{if }i\leq N\\
+    p_{i-N}&\quad\text{if }i> N
   \end{cases}
-  \qquad j=1, \ldots, 2N,
+  \qquad i=1, \ldots, 2N,
 \end{equation}
@@
 \begin{equation}\label{eq:std_ode}
-  \dot x_j=G_j(t, x_1, x_2, \ldots, x_{2N}),\quad j=1, \ldots, 2N
+  \dot \nu_i=G_i(t, \nu_1, \nu_2, \ldots, \nu_{2N}), \quad i=1, \ldots, 2N
 \end{equation}
@@
 \begin{equation}
-  G_j(t, x)=
-  \begin{cases}
-    \frac{\partial H}{\partial x_{j+N/2}},&\quad\text{if }j\leq N\\
-    \frac{\partial H}{\partial x_{j-N/2}},&\quad\text{if }j> N.
+  G_i(t, \nu)=
+  \begin{cases}\displaystyle
+    \pd{H}{\nu_{i+N}} ,&\quad\text{if }i\leq N\\[1em]
+    \displaystyle
+    -\pd{H}{\nu_{i-N}}, &\quad\text{if }i> N.
   \end{cases}
 \end{equation}
-This expression can be further simplified as a matrix product:
+This expression can be further shortened as a matrix product:
 \begin{equation}\label{eq:rhsmf}
-  G_i=M_{i j}\frac{\partial H}{\partial x_j},\qquad
+  G_i= \sum_j M_{i j}\pd{H}{\nu_j},\qquad
   \text{with }
@@
 system of $2N$ first order ordinary differential equations (ODEs) on $2N$
-variables $x_1, x_2, \ldots, x_{2N}$.
+variables $\nu_1, \nu_2, \ldots, \nu_{2N}$.
 
 In all but a small handfull of very well known simple problems, this system of
-ODEs must be solved numerically. All popular library routines for solving
-ODEs\footnote{Like ODEPACK \cite{odepack} for fortran, \texttt{odeint}
-  \cite{odeint} in c++ boost libraries or 
-  \texttt{scipy.integrate.solve\_ivp}~\cite{scipy} for python.}
-require the user to supply subprograms to compute the rhs functions $G_i$.
-The ODE solver invokes these subprograms to compute estimates for the
-values of the unknowns $x_i$ for arbitrary time $t$, given their values at a
-given instant $t_0$.
+ODEs has no analytical solutions and must be solved numerically.  Popular
+computer library routines for solving ODEs\footnote{Like ODEPACK~\cite{odepack}
+  for fortran, \texttt{odeint} \cite{odeint} in c++ boost libraries or
+\texttt{solve\_ivp} in SciPy~\cite{scipy} for Python.} require the user to
+supply subprograms to compute the right hand side functions $G_i$.  The ODE
+solver invokes these subprograms to compute estimates for the values of the
+unknowns $\nu_i$ for arbitrary time $t$, given their values at a particular
+instant $t_0$.
 
-In this work, we propose an alternative numerical method, where the user only
-supplies the hamiltonian function of the system, leaving its partial derivatives
-to be computed numerically as well.  This procedure is computationally more
-expensive, since many evaluations of the hamiltonian function are needed in
-order to estimate its partial derivatives at any moment in time, and so it
-really is not suitable for the analysis of complex systems with more than just a
-few coupled degreess of freedom. However, for simple systems it is very
-practical and, anyway, it is an interesting concept in its own, which we haven't
-seen exposed elsewhere.
-
-%-------------------------------------------------------------------------------
-\section{Numerical procedure}
-Standard numerical routines for solving systems of ODEs expect the problem to be
-written in the form
-\begin{equation}\label{eq:numheqs}
-  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
-\end{equation}
-where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
-equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
-G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
-the user as callable subprograms. For instance, the python standard initial
-value problem solver routine has interface
-\texttt{solve\_ivp(rhs, time\_interval, psi\_0)}; here, \texttt{rhs} is the name
-of a phython function to compute the values of the rhs functions $G_i$.
-During its execution, the ODE solver routine repeatedly invokes these
-subprograms (as many times as required by the specific integration algorithm
-and/or desired accuracy) in order to compute estimates for the values the
-unknown functions $\psi_i$ at given values of $t$.
-
-For Hamilton's equations, the $\psi_i$ are the coordinates and conjugate momenta
-of the system, while the $G_i$ are the partial derivatives of its hamiltonian
-function (reordered and eventually symmetrized, according to
-eq.~\ref{eq:rhsmf}). The usual procedure is to derive analytical expressions
-for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
-code those expressions as subprograms to be called by the ODE solver routine.
-
-In the method now proposed, instead, the user only has to supply code to
-compute the value of the system's hamiltonian function; its partial derivatives
-(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically and reordered by
-another subprogram which the ODE solver calls. Note that the differentiation and
-reordering of the partial derivatives can be computed by a general purpose
-routine. Indeed, the gradient can even be computed also by a standard numerical
-library function, if available for the chosen computing system or language.
+In this work, we propose an alternative numerical method (which we will refer to
+as \emph{Numerical Evolution from the Hamiltonian}, or NEVH) where the user only
+needs to write code for the Hamiltonian function of the system, leaving its
+partial derivatives to be computed numerically by a general purpose subprogram.
 Figure~\ref{fig:a} displays flowcharts for both methods.
@@
       \footnotesize
-      \tikzstyle{io} = [trapezium,
-                        trapezium left angle=70,
-                        trapezium right angle=110,
-                        minimum width=1.0cm,
-                        minimum height=.6cm,
-                        text centered,
-                        draw=black,
-                        fill=blue!20]
       \tikzstyle{libcode} = [rectangle,
@@
                              draw=black,
-                             fill=orange!20]
+                             %fill=orange!20
+                             ]
       \tikzstyle{usercode} = [rectangle,
@@
                              text width=1.5cm,
-                             fill=green!20]
+                             %fill=green!20
+                             ]
       \tikzstyle{arrow} = [thick,->,>=stealth]
@@
         \node at (0,1) {Standard method};
-        \node (input) [io] {Initial state};
+        \node (input) [usercode] {Initial state};
         \node (odesolver) [libcode, below of=input] {ODE Solver};
-        \node (output) [io, below of=odesolver] {Final state};
+        \node (output) [usercode, below of=odesolver] {Final state};
         \node (rhs) [usercode, right of=odesolver, node distance=2.3cm]
-          {$M\,\text{grad}\,H$};
+          {Subprogram for $G_i$};
         \draw [arrow] (input) -- (odesolver);
@@
       \begin{scope}[xshift=3.5cm]
-        \node at (0,1) {Proposed method};
-        \node (input) [io] {Initial state};
+        \node at (0,1) {NEVH};
+        \node (input) [usercode] {Initial state};
         \node (odesolver) [libcode, below of=input] {ODE Solver};
-        \node (output) [io, below of=odesolver] {Final state};
+        \node (output) [usercode, below of=odesolver] {Final state};
         \draw [arrow] (input) -- (odesolver);
         \draw [arrow] (odesolver) -- (output);
-        \node (rhs) [libcode, right of=odesolver, node distance=2.5cm]
-          {General purpose\\ $M\,\text{grad}\,f$ routine};
+        \node (rhs) [libcode,  right of=odesolver,
+                     node distance=2.5cm]
+                     {General purpose\\ $M_{ij}%\,
+                     \partial/\partial x_j$ subprogram};
         \draw [arrow] (odesolver) -- (rhs);
@@
 
-%To make this discussion clearer, let us consider a specific problem, the
-%classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
-%the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
-%a classical oscilator with mass $m$ and restoring force constant $k$ is 
-%\begin{equation}
-%  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
+This method is computationally more expensive the the standard approach, since
+at least two evaluations of the Hamiltonian function are needed to evaluate each
+of its partial derivatives at any moment in time, and so it really is not
+suitable for the analysis of complex systems with more than just a few coupled
+degrees of freedom. However, for simple systems it is very practical
+and, anyway, it is an interesting approach in its own, which we haven't seen
+exposed elsewhere.
+
+%-------------------------------------------------------------------------------
+\section{Numerical procedure and a simple implementation in Python}
+The ``Generic purpose $M_{ij}\partial/\partial x_j$ subprogram'' in the
+flowchart in Figure~\ref{fig:a} can be easily implemented in any modern computer
+language. Using a central difference formula for the partial derivatives, a
+pseudocode version can be sketched as
+\begin{algorithmic}[H]
+  \Function{G}{$H, t, \nu$}
+  \For{$1\leq i \leq 2N$}
+  \State
+  $\partial_iH\gets \frac{\displaystyle%
+  H(t, \tilde\nu_i,\nu_i+\delta\nu_i)-
+H(t, \tilde\nu_i,\nu_i-\delta\nu_i)}{\displaystyle2\delta\nu_i}
+  $
+  \EndFor\State
+  \Return $\{\partial_{N+1}H, \partial_{N+2}H, \ldots, \partial_{2N}H,
+  -\partial_1H, \partial_2H, \ldots, \partial_NH\}$
+  \EndFunction
+\end{algorithmic}
+(In this pseudocode definition, the notation $\tilde\nu_i$ stands for all the
+variables $\nu_1, \ldots, \nu_{2N}$, \emph{except} the $i$-th, $\nu_i$.)
+
+%Standard numerical routines for solving systems of ODEs expect the problem to be
+%written in the form
+%\begin{equation}\label{eq:numheqs}
+%  \frac{d\nu_i}{dt}=G_i(t, \nu_1, \nu_2, \ldots, \nu_{2N}),\quad i=1, \ldots, N,
 %\end{equation}
-%Hamilton's equations~\eqref{eq:heqs} then read
-%\begin{align}
-%  \frac{dq}{dt}&=\frac{p}{m}&
-%  \frac{dp}{dt}&=-kq,
-%\end{align}
-%wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
-%$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
-%by iterating the formulas
-%\begin{align}
-%  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
-%  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
-%\end{align}
-%where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
-%implements this method, and this the usual procedure. With this paper, we
-%consider a method where explicit expressions for the hamiltonian's partial
-%derivatives in the equations of motion are replaced by numerical estimates. In
-%our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
-%and~\eqref{eq:efe2} by
-%\begin{align*}
-%  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
-%  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
-%\end{align*}
-%where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
-%In order to make numerical aproximations to the partial derivatives of a
-%function, one must evaluate it several times; at the very least, two evaluations
-%are needed. For example, using the second order central difference method, the
-%derivative is estimated as
-%\begin{equation}\label{eq:cdiff}
-%  \frac{\partial\psi}{\partial x} = 
-%    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
-%  o(\delta x)^2,
-%\end{equation}
-%leading to approximation formulas that, for our example problem, read 
-%\begin{align}
-%  \numap{\frac{\partial H}{\partial p}}_i &=
-%  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
-%  \numap{\frac{\partial H}{\partial q}}_i &=
-%  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
-%\end{align}
-%where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
-%Other approximations have higher accuracy, at the cost of extra evaluations of
-%the function to be derived.
+%where $N$ is the number of unknown functions $\nu_i(t)$ (and the number of
+%equations in well posed problems) and $G_i(t, \nu_1, \nu_2, \ldots)\equiv
+%G_i(t,\nu_j)$ are the rhs of the differential equations, which must be coded by
+%the user as callable subprograms. For instance, the Python standard initial
+%value problem solver routine has interface
+%\texttt{solve\_ivp(rhs, time\_interval, psi\_0)}; here, \texttt{rhs} is the name
+%of a phython function to compute the values of the rhs functions $G_i$.
+%During its execution, the ODE solver routine repeatedly invokes these
+%subprograms (as many times as required by the specific integration algorithm
+%and/or desired accuracy) in order to compute estimates for the values the
+%unknown functions $\nu_i$ at given values of $t$.
+%
+%For Hamilton's equations, the $\nu_i$ are the coordinates and conjugate momenta
+%of the system, while the $G_i$ are the partial derivatives of its Hamiltonian
+%function (reordered and eventually multiplied by -1, according
+%to~(\ref{eq:rhsmf})). The usual procedure is to derive analytical expressions
+%for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
+%code those expressions as subprograms to be called by the ODE solver routine.
+%Quite often, this derivation is cumbersome, time consuming and error prone.
+%However, for simple systems at least, numerical calculation of the Hamiltonian
+%function may be a practical alternative.
+%
+%In the method now proposed, instead, the user only has to supply code to
+%compute the system's Hamiltonian; the rhs functions $G_i$ in~\eqref{eq:std_ode},
+%needed by the ODE solver, are computed numerically by a generic subroutine
+%which takes the Hamiltonian function as an argument and returns the values of
+%its partial derivatives, properly reordered and multiplied by $\pm1$ as
+%prescribed in~\eqref{eq:rhsmf}. This function is generic in the sense that it
+%can be written independently of the particular system under analysis or the
+%specific form of its Hamiltonian. For example, using central difference formulas
+%and keeping the notation of~\eqref{eq:xinot}, it can be defined as 
+%\begin{algorithmic}[H]
+%  \Function{G}{$H, t, \nu$}
+%  \For{$1\leq i \leq 2N$}
+%  \State
+%  $\partial_iH\gets \frac{\displaystyle%
+%  H(t, \tilde\nu,\nu_i+\delta\nu_i)-
+%H(t, \tilde\nu,\nu_i-\delta\nu_i)}{\displaystyle2\delta\nu_i}
+%  $
+%  \EndFor\State
+%  \Return $\{\partial_{N+1}H, \partial_{N+2}H, \ldots, \partial_{2N}H,
+%  -\partial_1H, \partial_2H, \ldots, \partial_NH\}$
+%  \EndFunction
+%\end{algorithmic}
+%(In this pseudocode definition, the notation $\tilde\nu$ stands for all the
+%variables $\nu_1, \ldots, \nu_{2N}$, except the $i$-th, $\nu_i$.)
+%In the method now proposed, instead, the user only has to supply code to
+%compute the value of the system's Hamiltonian function; its partial derivatives
+%(the $G_i$ in~\eqref{eq:numheqs}) are estimated numerically and reordered by
+%another subprogram which the ODE solver calls. Note that the differentiation and
+%reordering of the partial derivatives can be computed by a general purpose
+%routine. Indeed, the gradient can even be computed also by a standard numerical
+%library function, if available for the chosen computing system or language.
 
-%This need for repeated evaluation of the system's hamiltonian reduces the
-%efficiency of this method. However, programming a procedure for the hamiltonian
-%function is usually much simpler then doing it for the set of its partial
-%derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
-%hamiltonian procedure is likewise much simpler than when the partial derivatives
-%are coded.  For small systems, using the hamiltonian is much simpler and faster.
+We made a simple implementation of this method in Python, using the standard
+packages NumPy~\cite{numpy:2011,numpy:2020} and SciPy~\cite{scipy}.  It
+consists of a function that calls the Hamiltonian function supplied by
+the user to evaluate, reorder and return its partial derivatives, computed using
+second order central difference formulas.
 %
+%The step parameters $\delta\nu_i$ for
+%the derivation formulas must be supplied by the user.
+%This function is then
+%given
+%as an argument to SciPy's ODE solver \texttt{solve\_ivp}.
 %
-%
-\section{Simple implementation in Python and examples}
-We made a simple implementation of this method in Python, using the standard
-numerical extensions library NumPy~\cite{numpy:2011,numpy:2020}.  It consists of
-a simple function that calls the hamiltonian function supplied by the user to
-evaluate its partial derivatives, using the central difference formulas of
-eqs.~\eqref{eq:cdiff}. This function is then supplied to NumPy's ODE solver
-\texttt{solve\_ivp}.  For greater efficiency, it is wrapped in a class, whose
-objects store initialization details on creation, like the hamiltonian's
-parameters, the number of degrees of freedom, the values of the discretization
-steps, etc. These objects are callable with signature adjusted to what the ODE
-solver expects.  The whole class definition fits in a few dozen lines of code
-and is very easy to use. It is available for download as free software at
-github~\cite{nevh:2020}. The repository also includes a \texttt{jupyter
-notebook} showcasing several example applications.
+For greater simplicity of use, it is wrapped in a class (named \texttt{HGrad})
+whose objects store, on creation, all the problem physical and numerical
+details, like the Hamilton function and its parameters, the number of degrees of
+freedom, the values of the discretization steps $\delta\nu_i$, etc.
+\texttt{HGrad} objects are defined as callable with signature \texttt{objname(t,
+$\nu_i$)}, returning a list containing the values of the rhs functions $G_i$ in
+Hamilton's equations for given values of the dynamical variables of the
+dynamical variables $\nu_i$ at given time \texttt{t}. In this way, the calling
+syntax of \texttt{HGrad} objects is consistent with the interface of the ODE
+solver we chose for our implementation (SciPy's function \texttt{solve\_ivp})
+regarding the specification of the rhs $G_i$ functions, so that
+\texttt{HGrad} objects can be used
+just as if they were explicitly coded python functions.  The whole class
+definition (including comments and Python doc strings) fits in a few dozen lines
+of code and is very easy to use.  Our code is available for download as free
+software at github~\cite{nevh:2020}.  The repository also includes a
+\texttt{jupyter notebook} showcasing several example applications.
+
+Listing~\ref{lst:1dlho} presents a particularly simple application example, the
+one-dimensional harmonic oscillator. Most of the lines of code displayed deal
+with general adminis\-trative details (initialization of physical and numerical
+paramenters, definition of initial state, and so on), also needed for standard
+numerical resolutions of Hamilton's equations. In Listing~\ref{lst:1dlho}, the
+sections specific to our method are lines 6-8 (definition of the Hamiltonian
+function), line 13 (initialization of the discretization steps for the
+calculation of partial derivatives) and line 19 (definition of the
+\texttt{HGrad} object \texttt{G}).
+
+\begin{lstlisting}[caption= {Numerical solution of Hamiltonian's equations for
+the one-dimensional linear harmonic oscillator. Note that the user only supplies
+code for the system's Hamilton function in lines 7-9. The remaining code lines
+define initial state, physical and numerical parameters.},
+numbers=left, stepnumber=1, label={lst:1dlho}, language=python
+]
+from scipy.integrate import solve_ivp
+import nevh
+PI = 3.141592653589793
+
+# Hamiltonian function
+def H(t, s, k, m):
+    x, p = s
+    return 0.5 * k * x**2 + 0.5 * p**2 / m
+
+# Initial state: off equilibrium position, at rest (x=1, p=0)
+initial_state = [1.0, 0.0]
+# Discretization step parameters for numerical derivatives
+ds = [0.1, 0.1]
+# Hamiltonian parameters. With k=4\pi^2, m=1, the period is 1
+kc = 4 * PI**2
+mc = 1.0
+
+# Create the HGrad object, use scipy.solve_ivp to
+G = nevh.HGrad(H, ds, k=kc, m=mc)
+
+# solve Halmilton's eqs numerically from t=0 to t=3
+trj = solve_ivp(G, [0.0, 3.0], initial_state)
+\end{lstlisting}
+Note that the \texttt{HGrad} object \texttt{G} is supplied to scipy's ODE solver
+\texttt{solve\_ivp} just as if \texttt{G} was the name of a Python function for
+evaluating the rhs of Hamilton's equations.
+
+Scipy's ODE solver function \texttt{solve\_ivp} returns a composite structure
+(named \texttt{trj} in Listing~\ref{lst:1dlho}) that stores information
+on the numerical solution of the ODE, for example, sampled values of the
+coordinates and momenta at different times in the interval \texttt{[tmin,
+tmax]}.
+%are stored at the arrays \texttt{trj.y[0]}, \texttt{trj.y[1]}, while the
+%sampling times are stored in \texttt{trj.t}. The sampling frequency can be
+%adjusted by the user, but here we just took the default behaviour of
+%\texttt{solve\_ivp}, which sets that frequency accordind to precision
+%requirements (that we also didn't bother to specify, because it wasn't necessary
+%for convergence in this example).
+
+The values of the discretization steps used for the numerical computation of the
+partial derivatives of the Hamiltonian, are stored in array \texttt{ds} in
+Listing~\ref{lst:1dlho} (at line 13).  In the example shown these values are
+actually arbitrary, because the Hamiltonian of the linear harmonic oscilator is a
+quadratic function of all the coordinates and conjugate momenta and the central
+difference formulas yield exact estimates of the derivatives of quadratic
+functions. However, in other problems, the values of the discretizations steps
+may need to be more carefully considered.
+
+Figure~\ref{fig:1dlho} displays the values of the  coordinates and momenta of
+the numerical solution obtained (circles and squares, respectively) in arbitrary
+units, together with plots of the analytical solutions for the same problem
+(continuous and dashed lines).
+\begin{figure}[htb]
+  \begin{center}
+    \includegraphics[width=0.6\linewidth]{figs/shm.png}
+  \end{center}
+  \caption{Plot of position (\texttt{x}) and momentum (\texttt{p}) for the
+  one-dimensional harmonic oscillator problem of Listing~\ref{lst:1dlho}.
+  }\label{fig:1dlho}
+\end{figure}
+The correctness of the numerical approximation, at least for pedagogical
+purposes only, is manifest.
+
+The preceding example is so simple that the main advantage of our method is not
+really conveyed, since it's a trivial matter to derive and code the partial
+derivatives of the harmonic oscillator Hamiltonian. Let us now consider a more
+complex problem, the dynamics of the planar double pendulum. A planar double
+pendulum is a set of two simple pendulums where the second is suspended from the
+bob of the first (see Figure~\ref{fig:b}) and both are constrained to move in
+the same vertical plane. This system has two obvious degrees of freedom, the
+angles $\varphi_1$ and $\varphi_2$ which both pendulums define with the vertical
+(see Figure~\ref{fig:b}).
+\begin{figure}[htb]
+  {\centering
+    \begin{tikzpicture}
+      \small
+      \draw (-1,0) -- (1,0);
+      \foreach \x in {0.8, 0.6, 0.4, 0.2}
+      {
+        \draw (\x,0) -- +(60:0.3);
+        \draw (-\x,0) -- +(60:0.3);
+      }
+      \draw (0,0) -- (60:0.3);
+      \draw [thick] (0,0) -- (285:0.9) coordinate(a)
+        node[midway, right]{$l_1$} 
+        -- +(310:0.9) coordinate(b)
+        node[midway,right] {$l_2$};
+      \fill (a) circle(0.1) node [right]{$m_1$};
+      \fill (b) circle(0.1) node [right]{$m_2$};
+      \draw [dashed] (0,0) -- + (0,-1);
+      \draw (285:0.7) arc (285:270:0.7)node [left] {$\varphi_1$};
+      \draw [dashed] (a) --+ (0,-1);
+      \path (a) -- +(310:0.7) coordinate(c);
+      \draw (c) arc (310:270:0.7) node[left]{$\varphi_2$};
+    \end{tikzpicture}
+    \caption{\label{fig:b}Double pendulum.}
+  }\par
+\end{figure}
+
+For the case of equal masses $m$ and lengths $l$, the Hamiltonian
+reads~\cite{bohm:2018}
+\begin{equation}\label{eq:dph}
+  H=\frac{1}{2ml^2}
+  \frac{p_1^2 + p_2^2-2p_1p_2\cos(\varphi_1-\varphi_2)}%
+    {1+\sin^2(\varphi_1-\varphi_2)}-
+    mgl(2\cos\varphi_1+\cos\varphi_2),
+\end{equation}
+where the conjugate momenta $p_1$, $p_2$ are given by
+\begin{align}
+  p_1&=2ml^2\dot{\varphi_1}+ml^2\dot{\varphi_2}\cos(\varphi_1-\varphi_2)&
+  p_2&=2ml^2\dot{\varphi_2}+ml^2\dot{\varphi_1}\cos(\varphi_1-\varphi_2)
+\end{align}
+Calculating the partial derivatives of this Hamiltonian relative to the angles
+$\varphi_i$ and their conjugate momentum $p_i$ and programming the resulting
+expressions is quite tedious and error-prone. Here, NEVH is markedly
+simpler: the user just just needs to code the Hamiltonian~\eqref{eq:dph}, and
+that's all. No need to derive and code complicated expressions for the right
+hand side functions $G_i$ in Hamilton's equations~\eqref{eq:std_ode}.
+
+For the double pendulum, we implemented the two methods.  Both approaches yield
+similar results. Figure~\ref{fig:c} shows plots of the difference between
+results for the two angles $\varphi_1$ and $\varphi_2$ obtained using both
+methods, for the case $m=1$\,kg, $l=1$\,m, $g=9.8$\,m/s$^2$. The initial
+configuration is $\varphi_1=\pi/2,\ \varphi_2=0,\ p_1=p_2=0$.
+\begin{figure}[htb]
+  {\centering
+    \includegraphics[width=0.8\linewidth]{figs/dpchaos.png}
+    \caption{\label{fig:c}Double pendulum. The plots show the diference between
+    the angle values ($\varphi_1$ on the left, $\varphi_2$ to the right)
+    computed with the standard and the NEVH method. Initial state is
+    $\varphi_1=\pi/2$, $\varphi_2=0$, $p_1=p_2=0$. The ODE solver
+    (\texttt{scipy.integrate.solve\_ivp}) is configured similarly for both
+  methods.}
+  }\par
+\end{figure}
+Both methods initially yield essentially the same results. However, due to the
+well known chaotic nature of the double pendulum dynamics, small discrepancies
+grow up and sooner or later lead to completely different predicted evolutions of
+the system.  Expectedly, this amplifications of differences can be somewhat
+delayed by choosing smaller discretization steps in the numeric calculation of
+the partial derivatives of the Hamiltonian.
+
+In this example, the Hamiltonian function is five python lines long; the
+definition of the $G_i$ functions is twice as long, taking 12 lines of code. 
+It must also be said, however, that the NEVH code for this problem is around
+four times slower than the program for the standard approach.
+
+\section{Accuracy}
+The sources of numerical inaccuracy specific to the method here proposed are the
+errors in the numerical evaluation of the partial derivatives of the
+Hamiltonian. In our implementation, we used lowest order central difference
+formulas, and so these errors are expected to be of second order on the
+discretizations steps $\delta\nu_i$ (except for linear and quadratic functions,
+for which central difference formulas yield exact results).
+In order to check this, we compared results 
+ for a single
+time step evolution of a one-dimensional point partice subject to power law
+potentials, 
+with Hamiltonian function
+\begin{equation}\label{eq:errh}
+  H_{\alpha,n}(x,p)=\alpha x^n+\frac{1}{2m}p^2,
+\end{equation}
+for some real constant $\alpha$.
+We took as measure of error the $L_2$-norm of the difference of final states
+computed with NEVH and the standard approach.
+The results are plotted in the graph displayed in Figure~\ref{fig:e} as 
+functions of the discretization steps $\delta\nu_i$ (we
+took $\delta\nu_1=\delta\nu_2\equiv\delta\nu$ in arbitrary units).
+\begin{figure}[htb]
+  \centering
+    \includegraphics[width=0.6\linewidth]{figs/errs.png}
+  \par
+  \caption{\label{fig:e}L2-norm of the difference between results obtained
+    with the standard method and NEVH for a single time step (of duration
+    $\delta t=0.001$, in arbitrary units) evolution of a point particle with
+    hamiltonian \eqref{eq:errh} as a function of the discretizations steps
+    $\delta\nu_1=\delta\nu_2\equiv\delta\nu$. The values $\alpha=m=1$ (arbitrary
+    units) were used. The initial state was $x=1.0;\ p=0.0$.}
+\end{figure}
+
+The graph displays the expected $O(\delta\nu^2)$ behaviour, for
+$\delta\nu\gtrsim5\times10^{-6}$. For smaller values of $\delta\nu$, this tendency is
+not perceived because floating point round-off errors become dominant (at the
+floating point precision used in our calculation), hiding the contribution from
+the calculation of the partial derivatives. For the case $n=2$, the numerical
+derivation is exact and only round-off errors affect the results. These two
+facts agree with our expectations, supporting the conviction that NEVH is a
+well behaved, reliable method for the numerical solution of Hamilton's equations
+for simple, low-dimensional point particle systems.
 
 \section{Conclusion}
+For numerical simulations of the dynamics of simple point particle systems
+with few coupled degrees of freedom, NEVH is much simpler then the traditional
+approach, because the user is spared the trouble of deriving expressions for the
+partial derivatives of the Hamiltonian with respect to the system's coordinates
+and momenta and of writing subprograms to compute them. The user only needs to
+supply code for the system's Hamiltonian itself.
+
+It is true that NEVH runs
+slower than equivalent traditional approaches, because two or more evaluations
+of tht Hamiltonian are needed to compute each of its partical derivatives. 
+However, for simple systems, the time delay is hardly noticeable by the user.
+
+Regarding accuracy, NEVH behaves as expected and (also as expected) does not
+seem to introduce particular numerical instablities.
+
+For these reasons, we find quite reasonable that NEVH shoud be considered in
+numerical simulations of simple point particle systems, in particular for
+pedagogical purposes.
+
+%Note that we do not claim that this method is more efficient than the standard
+%approach. Much to the contrary, in computationally intensive applications, the
+%effort involved in the numerical calculation of the partial derivatives of the
+%Hamiltonian (for which several computations of the Hamiltonian function itself
+%are needed), at every instant in the simulation, slows the calculaton very
+%significantly. What we do claim is that our method makes it very simple to set
+%up a practical simulation of small mechanical systems, because the user is
+%spared the trouble of deriving expressions for the partial derivatives of the
+%Hamiltonian and writing subroutines that compute them, with the calling
+%signature that the particular ODE solver expects.
 
@@
   \texttt{https://github.com/ljmamoreira/nevh} (2020)
+%
+  \bibitem{bohm:2018} John L. Bohm, \textsl{A Student's Guide to Analytical
+  Mechanics.} Cambridge University Press (2018)
 

### 2022-09-07 src/grad.py

@@ -1,39 +1,0 @@
-# 2019-08-12
-# JML Amoreira: jmlamoreira@gmail.com
-# LJM Amoreira: amoreira@ubi.pt, ljmamoreira@gmail.com
-
-import numpy as np
-
-
-def grad(H, t, psi, dpsi, **hparams):
-    """grad(H, t, psi, dpsi, **hparams)
-    Given function H(t, psi, **hparams), return numerical estimate of gradient
-    at t, psi, using step vector dpsi.
-    The gradient is computed using central differences. The returned gradient
-    has shape equal to psi.shape. 
-
-    Parameters:
-    ===========
-    H: callable with signature H(t, psi, **params)
-    t: double
-    psi: array of doubles storing qs and ps
-    dpsi: array of doubles (steps for finite difference formulas)
-    **hparams: other parameters of H
-
-    Returns:
-    ========
-    gradient: ndarray
-    
-    Examples:
-    =========
-    >>> grad(lambda t,x: x[0]+x[1]**2, 0, [1,1] ,[0.1,0.1])
-    array([1., 2.])
-    """
-    n = len(psi)
-    identity = np.eye(n)
-    return np.array(
-            [(H(t, psi + dpsi[i]*identity[i]/2, **hparams)-
-              H(t, psi - dpsi[i]*identity[i]/2, **hparams))/dpsi[i]
-             for i in range(n)])
-
-

### 2022-09-07 src/nevh.py

@@ -8,8 +8,8 @@
 
 import numpy as np
 
-# Hgrad: Hamiltonian grandient. 
-class Hgrad():
+# HGrad: Hamiltonian grandient. 
+class HGrad():
     """
     A class to compute the hamiltonian gradient of hamiltonian funcs.
 
@@ -36,7 +36,7 @@ class Hgrad():
 
     def __init__(self, H, dpsi, **hparams):
         """
-        Creates and inits Hgrad objects
+        Creates and inits HGrad objects
 
         Parameters
         ----------
@@ -51,7 +51,7 @@ class Hgrad():
         # System global parameters  
         self.H = H
         self.N = len(dpsi) // 2     # Number of degrees of freedom
-        self.dpsi = dpsi            # Step sizes for differentiation
+        self.dpsi = dpsi              # Step fractions for differentiation
         self.hparams = hparams      # Hamiltonian parameters
         self.I = np.eye(2*self.N)   # Identity matrix
 

### 2022-09-07 tests/errs.py

@@ -1,0 +1,71 @@
+# 2021-10-31
+# amoreira@ubi.pt
+# Code for checking error dependence on derivation steps
+
+import sys
+import numpy as np
+import matplotlib.pyplot as plt
+sys.path.append("../src")
+import nevh
+
+PI = np.pi
+
+
+# Plain RK4
+def srk4(t, y, dt, f, *params):
+    k1 = f(t, y, *params)
+    k2 = f(t + dt/2, y + k1 * dt/2, *params)
+    k3 = f(t + dt/2, y + k2 * dt/2, *params)
+    k4 = f(t + dt, y + k3 * dt, *params)
+    return ((k1 + 2*k2+ 2*k3 + k4) * dt/6)
+
+
+# Power law potential hamiltonian
+# If exponent is negative, force is repulsive, but that's OK
+def plh(t, psi, m, gamma, n):
+    x, p = psi
+    return p**2/(2*m) - gamma * x**n
+
+
+# Power law potential Hamilton's eqs rhs
+def pleqrhs(t, psi, m, gamma, n):
+    x, p = psi
+    return np.array([p/m, n*gamma*x**(n-1)])
+
+
+# Compute errors for given n, given dx
+def err_n(n, dx):
+    m = 1.0
+    gamma = 1.0
+    psi_i = np.array([1.0, 0.0])
+    t = 0.0
+    dt = 0.05
+    dpsi_s = srk4(t, psi_i, dt, pleqrhs, m, gamma, n)
+
+    dqp = np.array([dx, 0.1], dtype=np.float32)
+    hgrd = nevh.HGrad(plh, dqp, m=m, gamma=gamma, n=n)
+    dpsi_n = srk4(t, psi_i, dt, hgrd)
+    err = np.linalg.norm(dpsi_n - dpsi_s)
+    return err
+
+
+if __name__ == "__main__":
+
+    dxs = [1e-7, 1e-6, 2e-6, 3e-6, 4e-6, 5e-6, 6e-6, 
+           7e-6, 8e-6, 9e-6, 1e-5, 1e-4, 1e-3, 1e-2]
+    ns = [2,3,4,5]
+    ls = [(),(3,3),(6,3),(9,3)]
+    dxs = np.logspace(-8,-1,400)
+    for n in ns:
+        errs = [] #np.zeros((len(ns),len(dxs)))
+        for dx in dxs:
+            err = err_n(n, dx)
+            #print(err)
[…]

### 2022-09-07 tests/halley.py

@@ -1,0 +1,97 @@
+import sys
+import numpy as np
+import matplotlib.pyplot as plt
+PI = np.pi
+sys.path.append("../src")
+import nevh
+
+def rk4pl(t, y, dt, f, *params):
+    k1 = f(t, y, *params)
+    k2 = f(t + dt/2, y + k1 * dt/2, *params)
+    k3 = f(t + dt/2, y + k2 * dt/2, *params)
+    k4 = f(t + dt, y + k3 * dt, *params)
+    return ((k1 + 2*k2+ 2*k3 + k4) * dt/6)
+
+
+# Hamiltonian (cartesian coordinates)
+def Hkep(t, psi, GM, mu):
+    x, y, px, py = psi
+    return (px**2 + py**2)/(2*mu) - GM*mu / (x**2 + y**2)**0.5
+
+
+# Hamilton's eqs rhs
+def heqsrhs(t, psi, GM, mu):
+    x, y, px, py = psi
+    r3 = (x**2 + y**2)**1.5
+    rhs = np.array([px/mu, py/mu, -GM*mu*x/r3, -GM*mu*y/r3])
+    return rhs
+
+
+def halley():
+    # Global physical parameters
+    GM = 39.426900 # AU^3/y^2
+    mu = 2.2       # 10^14 kg (Halley)
+    alpha = GM*mu
+
+    # Initial state (Halley's comet data)
+    xf = 35.082 #(apheliom)
+    xc = 0.586  #/periheliom
+    vf = (2*GM*(1/xc - 1/xf)/((xf/xc)**2-1))**0.5
+
+    # Angular momentum and energy
+    l = mu * xf * vf
+    E = mu * vf**2 / 2 - alpha / xf
+    # Eccentricity, semimajor axis
+    eps = (1 + 2*E*l**2/(mu*alpha**2))**0.5
+    a = (xc + xf) / 2
+    # Orbital period
+    T = (4*PI**2 * a**3/GM)**0.5
+
+    dt = 0.1
+    dpsi = np.array([0.2, 0.2, 0.1,0.1])
+    gkep = nevh.Hgrad(Hkep, dpsi, GM=GM, mu=mu)
+
+    psi_i = [xf, 0.0, 0.0, mu*vf]
+
+    psi_fn = rk4pl(0.0, psi_i, dt, gkep)
+    psi_fs = rk4pl(0.0, psi_i, dt, heqsrhs, GM, mu)
+    print(psi_i)
+    print(psi_fn)
+    print(psi_fs)
+
[…]
